---
title: "Synthea Healthcare Analytics"
subtitle: "Clinical Data Analysis — Patient Demographics, Anomaly Detection & Demand Forecasting"
author: "Abdulrahman Jalilov"
date: "`r format(Sys.Date(), '%B %d, %Y')`"
output:
  html_document:
    theme: flatly
    highlight: tango
    toc: true
    toc_float:
      smooth_scroll: true
    toc_depth: 3
    code_folding: hide
    df_print: paged
---

# Setup and Data Preparation

## Setup, Data Import, and Master Table

```{r setup-data-import-master,include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}

knitr::opts_chunk$set(echo=TRUE, warning=FALSE, message=FALSE)

setwd("C:/Users/aksta/Desktop")
options(encoding = "UTF-8")
Sys.setlocale("LC_ALL", "Turkish")

library(data.table)
library(dplyr)
library(ggplot2)
library(plotly)
library(lubridate)
library(scales)
library(knitr)
library(kableExtra)
library(reshape2)
library(corrplot)
library(survival)
library(stringr)

encounters = readRDS("C:/Users/aksta/Desktop/Synthea/encounters.rds")
observations  <- readRDS("C:/Users/aksta/Desktop/Synthea/observations .rds")
patients      <- readRDS("C:/Users/aksta/Desktop/Synthea/patients.rds")

dim(patients)
dim(encounters)
dim(observations)

head(patients, 3)
head(encounters, 3)
head(observations, 3)

colSums(is.na(patients))
colSums(is.na(encounters))
colSums(is.na(observations))

# Gender
patients[, .N, by = GENDER]

# Race
patients[, .N, by = RACE][order(-N)]


# Master cədvəlini də yenilə
patients[, age := as.integer(
  difftime(
    fifelse(is.na(DEATHDATE), as.IDate("2017-12-31"), DEATHDATE),
    BIRTHDATE, units = "days"
  ) / 365.25
)]

# =========================
# FIX: Dynamic age at each encounter
# Hər encounter-da xəstənin O GÜNKÜ yaşı hesablanır.
# Statik 2017 yaşı yalnız patient-level demografiya üçün saxlanılır.
# =========================
encounters[, encounter_date_tmp := as.IDate(DATE)]
encounters[patients, on = c(PATIENT = "ID"), BIRTHDATE_tmp := i.BIRTHDATE]
encounters[, age_at_encounter := as.integer(
  difftime(encounter_date_tmp, BIRTHDATE_tmp, units = "days") / 365.25
)]
encounters[, age_group_enc := fcase(
  age_at_encounter <= 21, "0-21",
  age_at_encounter <= 43, "22-43",
  age_at_encounter  < 65, "44-65",
  age_at_encounter >= 65, "65+",
  default = NA_character_
)]
encounters[, c("encounter_date_tmp","BIRTHDATE_tmp") := NULL]


# Encounter növləri
encounters[, .N, by = DESCRIPTION][order(-N)]
# Xəstə başına encounter sayı
enc_per_pt <- encounters[, .N, by = PATIENT]
summary(enc_per_pt$N)
observations[, value_num := suppressWarnings(as.numeric(VALUE))]

observations[
  DESCRIPTION %in% c(
    "Body Height", "Body Weight",
    "Systolic Blood Pressure", "Diastolic Blood Pressure",
    "Body Mass Index"
  ) & !is.na(value_num),
  .(
    N      = .N,
    Orta   = round(mean(value_num), 1),
    Median = round(median(value_num), 1),
    Min    = round(min(value_num), 1),
    Max    = round(max(value_num), 1)
  ),
  by = .(DESCRIPTION, UNITS)
]

# Encounter sayı hər xəstə üçün
enc_count <- encounters[, .(enc_count = .N), by = PATIENT]

# Hər xəstənin son vital ölçüləri
last_vitals <- observations[
  DESCRIPTION %in% c(
    "Body Mass Index",
    "Systolic Blood Pressure",
    "Diastolic Blood Pressure"
  ) & !is.na(value_num)
][order(PATIENT, DATE)][
  , .SD[.N], by = .(PATIENT, DESCRIPTION)
] |> dcast(PATIENT ~ DESCRIPTION, value.var = "value_num")

setnames(last_vitals,
  old = c("Body Mass Index", "Diastolic Blood Pressure", "Systolic Blood Pressure"),
  new = c("bmi", "dbp", "sbp")
)

# Master cədvəl
master <- patients |>
  merge(enc_count,   by.x = "ID", by.y = "PATIENT", all.x = TRUE) |>
  merge(last_vitals, by.x = "ID", by.y = "PATIENT", all.x = TRUE)

dim(master)
head(master[, .(ID, age, GENDER, RACE, enc_count, bmi, sbp, dbp)], 5)
library(stringr)
master <- master %>%
  mutate(
    FIRST = str_remove_all(FIRST, "\\d+"),
    LAST = str_remove_all(LAST, "\\d+"),
    MAIDEN = str_remove_all(MAIDEN, "\\d+")
  )

master[, age := patients[match(master$ID, patients$ID), age]]

summary(master$age)
```


# Descriptive Analysis

## Gender Distribution


```{r descriptive-gender-distribution, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}

gender_summary <- master[
  !is.na(GENDER),
  .N,
  by = GENDER
][order(-N)]

gender_summary[, percent := round(N/sum(N)*100,1)]

plot_ly(
  data = gender_summary,
  labels = ~GENDER,
  values = ~N,
  type = "pie",
  hole = 0.55,
  textinfo = "label+percent",
  textposition = "outside",
  hovertemplate =
    "<b>%{label}</b><br>"  %+%
    "Say: %{value}<br>"    %+%
    "Faiz: %{percent}<extra></extra>",
  marker = list(
    colors = c("#5E60CE","#64DFDF","#80FFDB"),
    line = list(color = "white", width = 3)
  )
) |>
  layout(
    title = list(
      text = "Gender Distribution",
      x = 0.5,
      font = list(size = 22)
    ),
    showlegend = TRUE,
    annotations = list(
      list(
        text = paste0(sum(gender_summary$N), "<br>Patients"),
        x = 0.5,
        y = 0.5,
        showarrow = FALSE,
        font = list(size = 18)
      )
    )
  )

```

## Age Group Distribution

```{r descriptive-age-group-distribution, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}
master<- master %>%  
  mutate(
    age_group = case_when(
      age >= 0  & age <= 21 ~ "0-21",
      age >= 22 & age <= 43 ~ "22-43",
      age >= 44 & age < 65  ~ "44-65",
      age >= 65             ~ "65+",
      TRUE                  ~ NA_character_
    )
  )
age_group_summary <- master[
  !is.na(age_group),
  .N,
  by = age_group
][order(age_group)]

p_age <- ggplot(age_group_summary,
       aes(x = age_group,
           y = N,
           fill = age_group)) +
  
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  
  geom_text(
    aes(label = scales::comma(N)),
    vjust = -0.3,
    size = 4
  ) +
  
  labs(
    title = "Age Group Distribution",
    x = "Age Groups",
    y = "Patient Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    legend.position = "none",
    
    axis.text.x = element_text(
      face = "bold",
      size = 11
    ),
    
    axis.text.y = element_text(
      size = 10
    )
  )

p_age

ggplotly(p_age)
```

## Top 10 Encounter Types

```{r descriptive-top-10-encounter-types, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}

enc_top10 <- encounters[
  !is.na(DESCRIPTION),
  .N,
  by = DESCRIPTION
][order(-N)][1:10]

p_enc <- ggplot(enc_top10,
       aes(x = reorder(DESCRIPTION, N),
           y = N,
           fill = N)) +
  
  geom_col(
    width = 0.75,
    color = "white"
  ) +
  
  geom_text(
    aes(label = scales::comma(N)),
    hjust = -0.1,
    size = 3.8
  ) +
  
  coord_flip() +
  
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  
  labs(
    title = "Top 10 Encounter Types",
    x = "",
    y = "Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    
    legend.position = "none",
    
    axis.text.y = element_text(
      face = "bold",
      size = 10
    )
  )

p_enc

ggplotly(p_enc)


```
## Encounter Type Intensity by Age Group

```{r descriptive-encounter-type-intensity-age, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}



# Encounter + patient age group
enc_age_data <- encounters[
  master,
  on = c(PATIENT = "ID")
]

# Top 10 encounter type
top_enc_types <- enc_age_data[
  !is.na(DESCRIPTION) & !is.na(age_group),
  .N,
  by = DESCRIPTION
][order(-N)][1:10, DESCRIPTION]

# Age group + encounter type count
enc_age_summary <- enc_age_data[
  DESCRIPTION %in% top_enc_types &
    !is.na(age_group),
  .N,
  by = .(age_group, DESCRIPTION)
]

# Age group order
enc_age_summary[, age_group := factor(
  age_group,
  levels = c("0-21", "22-43", "44-65", "65+")
)]

# Encounter type order
enc_age_summary[, DESCRIPTION := reorder(DESCRIPTION, N)]

# Bubble chart
p_enc_age_bubble <- ggplot(
  enc_age_summary,
  aes(
    x = age_group,
    y = DESCRIPTION,
    size = N,
    fill = N,
    text = paste0(
      "Age Group: ", age_group,
      "<br>Encounter Type: ", DESCRIPTION,
      "<br>Count: ", scales::comma(N)
    )
  )
) +
  geom_point(
    shape = 21,
    color = "white",
    alpha = 0.85
  ) +
  scale_size_continuous(
    range = c(4, 22),
    labels = scales::comma
  ) +
  scale_fill_gradient(
    low = "#BDE0FE",
    high = "#023E8A",
    labels = scales::comma
  ) +
  labs(
    title = "Encounter Type Intensity by Age Group",
    subtitle = "Bubble size shows encounter count for each age group",
    x = "Age Group",
    y = "Encounter Type",
    size = "Count",
    fill = "Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 11
    ),
    axis.text.x = element_text(
      face = "bold",
      size = 11
    ),
    axis.text.y = element_text(
      face = "bold",
      size = 9
    ),
    legend.position = "right"
  )

p_enc_age_bubble

ggplotly(p_enc_age_bubble, tooltip = "text")

```



## Clinical Indicator Measurement Counts

>Terminologiyalar qısa izah
Body Mass Index (BMI)
Boy və çəkiyə əsasən bədən çəkisinin normal olub-olmadığını göstərir.
Systolic Blood Pressure
Ürək qanı pompalayanda yaranan “yuxarı təzyiq” göstəricisidir.
Diastolic Blood Pressure
Ürək istirahətdə olarkən damar təzyiqini göstərən “aşağı təzyiq” göstəricisidir.
Hemoglobin A1c (HbA1c)
Son 2-3 ay üzrə orta qan şəkəri səviyyəsini göstərir və diabet monitorinqində istifadə olunur.
Glucose
Qanda şəkər səviyyəsini ölçür.
Triglycerides
Qandakı yağ səviyyəsini göstərir və ürək-damar riskləri ilə əlaqəlidir.
Cholesterol
Qandakı xolesterol səviyyəsini göstərir və kardiovaskulyar risk analizi üçün vacibdir

```{r descriptive-clinical-indicator-counts, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}

clinical_counts <- observations[
  DESCRIPTION %in% c(
    "Total Cholesterol",
    "Triglycerides",
    "Glucose",
    "Hemoglobin A1c/Hemoglobin.total in Blood",
    "Body Mass Index",
    "Diastolic Blood Pressure",
    "Systolic Blood Pressure"
  ) & !is.na(value_num),
  .N,
  by = DESCRIPTION
]

clinical_counts[, short_name := case_when(
  DESCRIPTION == "Total Cholesterol" ~ "Total Chol",
  DESCRIPTION == "Triglycerides" ~ "Triglycerides",
  DESCRIPTION == "Glucose" ~ "Glucose",
  DESCRIPTION == "Hemoglobin A1c/Hemoglobin.total in Blood" ~ "HbA1c",
  DESCRIPTION == "Body Mass Index" ~ "BMI",
  DESCRIPTION == "Diastolic Blood Pressure" ~ "Diastolic BP",
  DESCRIPTION == "Systolic Blood Pressure" ~ "Systolic BP",
  TRUE ~ DESCRIPTION
)]

clinical_counts[, short_name := factor(
  short_name,
  levels = c("Total Chol", "Triglycerides", "Glucose",
             "HbA1c", "BMI", "Diastolic BP", "Systolic BP")
)]

 ggplot(
  clinical_counts,
  aes(x = short_name, y = N, fill = short_name)
) +
  geom_col(width = 0.65, color = "white") +
  geom_text(
    aes(label = scales::comma(N)),
    vjust = -0.4,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Əsas klinik göstəriciləri üzrə ölçmə sayı",
    x = "",
    y = "Ölçmə sayı",
    fill = ""
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    legend.position = "top",
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10)
  )


```
## Detailed Age Distribution

```{r descriptive-detailed-age-distribution, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}


age_data <- master[, .N, by = age][order(age)]

plot_ly(
  data = age_data,
  x    = ~age,
  y    = ~N,
  type = "bar",
  marker = list(color = "#7F77DD")
) |>
  layout(
    title  = "Xəstələrin yaş paylanması",
    xaxis  = list(title = "Yaş"),
    yaxis  = list(title = "Xəstə sayı"),
    hovermode = "x"
  )

```

## Hospital Usage Intensity

```{r descriptive-hospital-usage-intensity, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}

enc_usage <- master[, .(
  encounter_group = case_when(
    enc_count == 0 ~ "0",
    enc_count >= 1 & enc_count <= 5 ~ "1-5",
    enc_count >= 6 & enc_count <= 15 ~ "6-15",
    enc_count >= 16 & enc_count <= 30 ~ "16-30",
    enc_count > 30 ~ "30+"
  )
)][
  , .N, by = encounter_group
]

p_usage <- ggplot(enc_usage,
       aes(x = encounter_group,
           y = N,
           fill = encounter_group)) +
  
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  
  geom_text(
    aes(label = scales::comma(N)),
    vjust = -0.3,
    size = 4
  ) +
  
  labs(
    title = "Hospital Usage Intensity",
    x = "Encounter Groups",
    y = "Patient Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    legend.position = "none"
  )

p_usage

ggplotly(p_usage)

```


## Average BMI by Age Group

```{r descriptive-average-bmi-by-age-group, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}

# =========================
# FIX: Encounter-level BMI (teacher fix applied)
# Hər encounter-da xəstənin O GÜNƏ ən yaxın BMI ölçüsü götürülür.
# Bütün illərin ortalaması deyil — bu tibbi cəhətdən mənasızdır.
# =========================

obs_bmi_tmp <- observations[
  DESCRIPTION == "Body Mass Index" & !is.na(value_num),
  .(PATIENT, obs_date = as.IDate(DATE), bmi = value_num)
]
setkey(obs_bmi_tmp, PATIENT, obs_date)

enc_tmp <- encounters[!is.na(age_group_enc),
  .(PATIENT, enc_date = as.IDate(DATE), age_group_enc)]
setkey(enc_tmp, PATIENT, enc_date)

enc_bmi_joined <- obs_bmi_tmp[enc_tmp,
  on = .(PATIENT, obs_date <= enc_date),
  mult = "last",
  .(PATIENT, enc_date, age_group_enc, bmi = x.bmi)
]

master[, age_group2 := cut(
  age,
  breaks = c(0,10,20,30,40,50,60,70,80,90,Inf),
  labels = c("0-9","10-19","20-29","30-39","40-49",
             "50-59","60-69","70-79","80-89","90+"),
  right = FALSE
)]

bmi_age_avg <- enc_bmi_joined[
  !is.na(age_group_enc) & !is.na(bmi),
  .(avg_bmi = round(mean(bmi, na.rm = TRUE), 1)),
  by = .(age_group2 = age_group_enc)
]

p_bmi_avg <- ggplot(
  bmi_age_avg,
  aes(x = age_group2,
      y = avg_bmi,
      fill = avg_bmi)
) +
  
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  
  geom_text(
    aes(label = avg_bmi),
    vjust = -0.4,
    size = 4
  ) +
  
  labs(
    title = "Average BMI by Age Group",
    x = "Age Group",
    y = "Average BMI"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    
    legend.position = "none",
    
    axis.text.x = element_text(
      face = "bold"
    )
  )

p_bmi_avg

ggplotly(p_bmi_avg)
```

## Average Systolic Blood Pressure by Age Group

```{r descriptive-average-systolic-bp-by-age, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}

# =========================
# FIX: Encounter-level SBP (teacher fix applied)
# =========================
obs_sbp_tmp <- observations[
  DESCRIPTION == "Systolic Blood Pressure" & !is.na(value_num),
  .(PATIENT, obs_date = as.IDate(DATE), sbp = value_num)
]
setkey(obs_sbp_tmp, PATIENT, obs_date)

enc_sbp_joined <- obs_sbp_tmp[enc_tmp,
  on = .(PATIENT, obs_date <= enc_date),
  mult = "last",
  .(PATIENT, enc_date, age_group_enc, sbp = x.sbp)
]

sbp_age_avg <- enc_sbp_joined[
  !is.na(age_group_enc) & !is.na(sbp),
  .(avg_sbp = round(mean(sbp, na.rm = TRUE), 1)),
  by = .(age_group2 = age_group_enc)
]

p_sbp_avg <- ggplot(
  sbp_age_avg,
  aes(x = age_group2,
      y = avg_sbp,
      fill = avg_sbp)
) +
  
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  
  geom_text(
    aes(label = avg_sbp),
    vjust = -0.4,
    size = 4
  ) +
  
  labs(
    title = "Average Systolic BP by Age Group",
    x = "Age Group",
    y = "Average Systolic BP"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    
    legend.position = "none",
    
    axis.text.x = element_text(
      face = "bold"
    )
  )

p_sbp_avg

ggplotly(p_sbp_avg)


```



## Average Systolic Blood Pressure by BMI Category

```{r descriptive-systolic-bp-by-bmi-category, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}
master <- as.data.table(master)

# Lazımi column-lar numeric olsun
master[, bmi := as.numeric(bmi)]
master[, sbp := as.numeric(sbp)]

# Əgər bmi_category yoxdursa, yaradırıq
master[, bmi_category := case_when(
  !is.na(bmi) & bmi < 18.5 ~ "Underweight",
  !is.na(bmi) & bmi >= 18.5 & bmi < 25 ~ "Normal",
  !is.na(bmi) & bmi >= 25 & bmi < 30 ~ "Overweight",
  !is.na(bmi) & bmi >= 30 ~ "Obese",
  TRUE ~ "Unknown"
)]

# BMI category order
master[, bmi_category := factor(
  bmi_category,
  levels = c("Underweight", "Normal", "Overweight", "Obese", "Unknown")
)]

# Average Systolic BP by BMI Category
sbp_bmi_cat <- master[
  !is.na(bmi_category) &
    bmi_category != "Unknown" &
    !is.na(sbp),
  .(
    avg_sbp = round(mean(sbp, na.rm = TRUE), 1),
    patient_count = .N
  ),
  by = bmi_category
]

sbp_bmi_cat[, bmi_category := factor(
  bmi_category,
  levels = c("Underweight", "Normal", "Overweight", "Obese")
)]

sbp_bmi_cat <- sbp_bmi_cat[order(bmi_category)]

sbp_bmi_cat

# Chart
p_sbp_bmi <- ggplot(
  sbp_bmi_cat,
  aes(
    x = bmi_category,
    y = avg_sbp,
    fill = bmi_category,
    text = paste0(
      "BMI Category: ", bmi_category,
      "<br>Average Systolic BP: ", avg_sbp,
      "<br>Patient Count: ", scales::comma(patient_count)
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = paste0(avg_sbp, "\n(n=", scales::comma(patient_count), ")")),
    vjust = -0.3,
    size = 4,
    fontface = "bold"
  ) +
  labs(
    title = "Average Systolic BP by BMI Category",
    x = "BMI Category",
    y = "Average Systolic BP"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    legend.position = "none",
    axis.text.x = element_text(face = "bold", size = 11)
  )

p_sbp_bmi

ggplotly(p_sbp_bmi, tooltip = "text")

```
## Average Total Cholesterol by Glucose Category

```{r descriptive-cholesterol-by-glucose-category, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

observations <- as.data.table(observations)
master <- as.data.table(master)

# VALUE numeric olsun
observations[, value_num := suppressWarnings(as.numeric(VALUE))]

# 1. Glucose və Total Cholesterol son dəyərləri

glucose_chol_data <- observations[
  DESCRIPTION %in% c(
    "Glucose",
    "Total Cholesterol"
  ) &
    !is.na(value_num),
  .(
    PATIENT,
    DATE,
    DESCRIPTION,
    value_num
  )
][
  order(PATIENT, DESCRIPTION, DATE)
][
  ,
  .SD[.N],
  by = .(PATIENT, DESCRIPTION)
]

glucose_chol_wide <- dcast(
  glucose_chol_data,
  PATIENT ~ DESCRIPTION,
  value.var = "value_num"
)

setDT(glucose_chol_wide)

setnames(
  glucose_chol_wide,
  old = c("Glucose", "Total Cholesterol"),
  new = c("glucose", "cholesterol"),
  skip_absent = TRUE
)

# Əgər master-da köhnə glucose/cholesterol varsa, silirik
old_cols <- intersect(
  names(master),
  c("glucose", "cholesterol", "glucose_category")
)

if (length(old_cols) > 0) {
  master[, (old_cols) := NULL]
}

# Master ilə merge
master <- merge(
  master,
  glucose_chol_wide,
  by.x = "ID",
  by.y = "PATIENT",
  all.x = TRUE
)

master[, glucose := as.numeric(glucose)]
master[, cholesterol := as.numeric(cholesterol)]

# 2. Glucose category

master[, glucose_category := case_when(
  glucose < 70 ~ "Low",
  glucose >= 70 & glucose < 100 ~ "Normal",
  glucose >= 100 & glucose < 126 ~ "Prediabetic",
  glucose >= 126 ~ "Diabetic",
  TRUE ~ "Unknown"
)]

master[, glucose_category := factor(
  glucose_category,
  levels = c("Low", "Normal", "Prediabetic", "Diabetic", "Unknown")
)]

# 3. Average cholesterol by glucose category


chol_by_glucose <- master[
  glucose_category != "Unknown" &
    !is.na(cholesterol),
  .(
    patient_count = .N,
    avg_cholesterol = round(mean(cholesterol, na.rm = TRUE), 1),
    median_cholesterol = round(median(cholesterol, na.rm = TRUE), 1),
    min_cholesterol = min(cholesterol, na.rm = TRUE),
    max_cholesterol = max(cholesterol, na.rm = TRUE)
  ),
  by = glucose_category
][order(glucose_category)]

# 4. Visual

p_chol_glucose <- ggplot(
  chol_by_glucose,
  aes(
    x = glucose_category,
    y = avg_cholesterol,
    fill = glucose_category,
    text = paste0(
      "Glucose Category: ", glucose_category,
      "<br>Average Cholesterol: ", avg_cholesterol,
      "<br>Median Cholesterol: ", median_cholesterol,
      "<br>Patient Count: ", scales::comma(patient_count),
      "<br>Min Cholesterol: ", min_cholesterol,
      "<br>Max Cholesterol: ", max_cholesterol
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = paste0(
      avg_cholesterol,
      "\n(n=", scales::comma(patient_count), ")"
    )),
    vjust = -0.3,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Average Total Cholesterol by Glucose Category",
    subtitle = "Descriptive comparison of cholesterol levels across glucose groups",
    x = "Glucose Category",
    y = "Average Total Cholesterol"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold", size = 10),
    legend.position = "none"
  )

p_chol_glucose

ggplotly(p_chol_glucose, tooltip = "text")
```


## Patient Monitoring Level

```{r descriptive-patient-monitoring-level, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}
monitoring <- observations[
  DESCRIPTION %in% c(
    "Body Mass Index",
    "Systolic Blood Pressure",
    "Diastolic Blood Pressure",
    "Glucose",
    "Total Cholesterol",
    "Triglycerides",
    "Hemoglobin A1c/Hemoglobin.total in Blood"
  ),
  .N,
  by = PATIENT
]

setnames(
  monitoring,
  "N",
  "measurement_count"
)

master <- merge(
  master,
  monitoring,
  by.x = "ID",
  by.y = "PATIENT",
  all.x = TRUE
)

master[, monitoring_level := case_when(
  measurement_count <= 5 ~ "Low Monitoring",
  measurement_count > 5 & measurement_count <= 20 ~ "Moderate Monitoring",
  measurement_count > 20 ~ "High Monitoring"
)]

monitor_summary <- master[
  ,
  .N,
  by = monitoring_level
]

monitor_summary[, monitoring_level := factor(
  monitoring_level,
  levels = c(
    "Low Monitoring",
    "Moderate Monitoring",
    "High Monitoring"
  )
)]

p_monitor <- ggplot(
  monitor_summary,
  aes(x = monitoring_level,
      y = N,
      fill = monitoring_level)
) +
  
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  
  geom_text(
    aes(label = scales::comma(N)),
    vjust = -0.3,
    size = 4
  ) +
  
  labs(
    title = "Patient Monitoring Level",
    x = "",
    y = "Patient Count"
  ) +
  
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.15))
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    
    legend.position = "none"
  )

p_monitor

ggplotly(p_monitor)


```


## Top 10 Reasons for Hospital Visits
```{r descriptive-top-10-visit-reasons, include=TRUE,message=FALSE,warning=FALSE,echo=TRUE}
visit_reason <- encounters[
  !is.na(REASONDESCRIPTION),
  .N,
  by = REASONDESCRIPTION
][order(-N)][1:10]

p_reason <- ggplot(
  visit_reason,
  aes(x = reorder(REASONDESCRIPTION, N),
      y = N,
      fill = N)
) +
  
  geom_col(
    width = 0.75,
    color = "white"
  ) +
  
  geom_text(
    aes(label = scales::comma(N)),
    hjust = -0.1,
    size = 3.8
  ) +
  
  coord_flip() +
  
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  
  labs(
    title = "Top 10 Reasons for Hospital Visits",
    x = "",
    y = "Encounter Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    
    legend.position = "none",
    
    axis.text.y = element_text(
      face = "bold",
      size = 9
    )
  )

p_reason



```


# Anomaly Analysis

## Observation Before Encounter Date Gap Anomaly

```{r anomaly-observation-before-encounter, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(ggplot2)
library(plotly)
library(scales)

observations <- as.data.table(observations)
encounters <- as.data.table(encounters)

# Tarixləri düzəldirik
observations[, obs_date := as.IDate(DATE)]
encounters[, encounter_date := as.IDate(DATE)]

# Observation + Encounter join
obs_enc_date <- observations[
  encounters[, .(
    ID,
    encounter_date,
    encounter_type = DESCRIPTION,
    reason_description = REASONDESCRIPTION
  )],
  on = c(ENCOUNTER = "ID"),
  nomatch = 0
]

# Gün fərqi: positive = observation encounter-dən əvvəldir
obs_enc_date[, days_before_encounter := as.integer(encounter_date - obs_date)]

# Sadəcə observation encounter-dən əvvəl olanlar
obs_before_encounter <- obs_enc_date[
  !is.na(obs_date) &
    !is.na(encounter_date) &
    obs_date < encounter_date
]

# Gap category
obs_before_encounter[, gap_group := case_when(
  days_before_encounter >= 1 & days_before_encounter <= 1 ~ "0-1 day before",
  days_before_encounter >= 2 & days_before_encounter <= 7 ~ "2-7 days before",
  days_before_encounter >= 8 & days_before_encounter <= 30 ~ "8-30 days before",
  days_before_encounter >= 31 & days_before_encounter <= 365 ~ "31-365 days before",
  days_before_encounter > 365 ~ "1+ year before",
  TRUE ~ "Other"
)]

obs_before_encounter[, gap_group := factor(
  gap_group,
  levels = c(
    "0-1 day before",
    "2-7 days before",
    "8-30 days before",
    "31-365 days before",
    "1+ year before"
  )
)]

# Summary
obs_before_summary <- obs_before_encounter[
  ,
  .(
    record_count = .N,
    unique_patients = uniqueN(PATIENT),
    min_days_before = min(days_before_encounter, na.rm = TRUE),
    max_days_before = max(days_before_encounter, na.rm = TRUE)
  ),
  by = gap_group
][order(gap_group)]

obs_before_summary

# Daha ciddi anomaly: 31+ gün əvvəl olanlar
obs_before_anomaly <- obs_before_encounter[
  days_before_encounter >= 31
]

obs_before_anomaly_summary <- obs_before_anomaly[
  ,
  .(
    anomaly_records = .N,
    unique_patients = uniqueN(PATIENT),
    max_days_before = max(days_before_encounter, na.rm = TRUE)
  )
]

obs_before_anomaly_summary

# Description üzrə hansı testlərdə daha çoxdur?
obs_before_by_description <- obs_before_anomaly[
  ,
  .(
    anomaly_records = .N,
    unique_patients = uniqueN(PATIENT),
    avg_days_before = round(mean(days_before_encounter, na.rm = TRUE), 1),
    max_days_before = max(days_before_encounter, na.rm = TRUE)
  ),
  by = DESCRIPTION
][order(-anomaly_records)][1:15]

obs_before_by_description

# Visual
p_obs_before <- ggplot(
  obs_before_summary,
  aes(
    x = gap_group,
    y = record_count,
    fill = gap_group,
    text = paste0(
      "Gap Group: ", gap_group,
      "<br>Record Count: ", scales::comma(record_count),
      "<br>Unique Patients: ", scales::comma(unique_patients),
      "<br>Min Days Before: ", min_days_before,
      "<br>Max Days Before: ", max_days_before
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = scales::comma(record_count)),
    vjust = -0.3,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Observation Before Encounter Date Gap",
    subtitle = "Large pre-encounter gaps were reviewed as timeline anomalies",
    x = "Time Gap",
    y = "Record Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(angle = 25, hjust = 1, face = "bold"),
    legend.position = "none"
  )

p_obs_before

ggplotly(p_obs_before, tooltip = "text")
```
## Death Certification Before Death Anomaly
```{r anomaly-death-certification-before-death, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(ggplot2)
library(plotly)
library(scales)

patients <- as.data.table(patients)
encounters <- as.data.table(encounters)

# =========================
# 1. Date preparation
# =========================

patients[, death_date := as.IDate(DEATHDATE)]
encounters[, encounter_date := as.IDate(DATE)]

# =========================
# 2. Death Certification encounter-ləri tapırıq
# =========================

death_cert_encounters <- encounters[
  grepl("Death Certification", DESCRIPTION, ignore.case = TRUE),
  .(
    encounter_id = ID,
    PATIENT,
    encounter_date,
    encounter_type = DESCRIPTION,
    reason_description = REASONDESCRIPTION
  )
]

# Əgər Death Certification tapılmasa, mesaj versin
if (nrow(death_cert_encounters) == 0) {
  
  print("Death Certification encounter tapılmadı. DESCRIPTION dəyərlərini yoxla.")
  
} else {
  
  # =========================
  # 3. Patient death date və ad-soyad əlavə edirik
  # =========================
  
  death_cert_with_patient <- merge(
    death_cert_encounters,
    patients[
      ,
      .(
        PATIENT = ID,
        death_date,
        FIRST,
        LAST,
        GENDER,
        RACE
      )
    ],
    by = "PATIENT",
    all.x = TRUE
  )
  
  # Ad-soyad təmizliyi
  death_cert_with_patient[, FIRST := gsub("[0-9]", "", FIRST)]
  death_cert_with_patient[, LAST := gsub("[0-9]", "", LAST)]
  
  death_cert_with_patient[, patient_name := trimws(paste(FIRST, LAST))]
  
  death_cert_with_patient[
    is.na(patient_name) |
      patient_name == "" |
      patient_name == "NA NA",
    patient_name := "Unknown Patient"
  ]
  
  # =========================
  # 4. Death Certification ölüm tarixindən əvvəl olubmu?
  # =========================
  
  death_cert_with_patient[, days_before_death := as.integer(death_date - encounter_date)]
  
  death_cert_before_death <- death_cert_with_patient[
    !is.na(death_date) &
      !is.na(encounter_date) &
      encounter_date < death_date
  ][order(-days_before_death)]
  
  death_cert_before_death
  
  # =========================
  # 5. Əgər anomaly yoxdursa mesaj versin
  # =========================
  
  if (nrow(death_cert_before_death) == 0) {
    
    print("Death Certification before death date anomaly tapılmadı.")
    
  } else {
    
    # Gün fərqini ilə çeviririk
    death_cert_before_death[, years_before_death := round(days_before_death / 365.25, 1)]
    
    # Top 10 ən böyük fərq
    top_death_cert_years <- death_cert_before_death[
      order(-years_before_death)
    ][1:min(10, .N)]
    
    # Label
    top_death_cert_years[, patient_label := paste0(
      patient_name,
      " (",
      substr(PATIENT, 1, 6),
      "...)"
    )]
    
    top_death_cert_years[, patient_label := factor(
      patient_label,
      levels = rev(patient_label)
    )]
    
    # =========================
    # 6. Chart
    # =========================
    
    p_death_years <- ggplot(
      top_death_cert_years,
      aes(
        y = patient_label,
        x = years_before_death,
        text = paste0(
          "Patient: ", patient_name,
          "<br>Patient ID: ", PATIENT,
          "<br>Certification Date: ", encounter_date,
          "<br>Death Date: ", death_date,
          "<br>Gap: ", years_before_death, " years",
          "<br>Days Before Death: ", scales::comma(days_before_death)
        )
      )
    ) +
      geom_segment(
        aes(
          x = 0,
          xend = years_before_death,
          y = patient_label,
          yend = patient_label
        ),
        color = "gray70",
        linewidth = 1
      ) +
      geom_point(
        size = 4,
        color = "#2C7FB8"
      ) +
      geom_text(
        aes(label = paste0(years_before_death, " yr")),
        hjust = -0.15,
        size = 4,
        fontface = "bold"
      ) +
      scale_x_continuous(
        labels = function(x) paste0(x, " yr"),
        expand = expansion(mult = c(0, 0.20))
      ) +
      labs(
        title = "Top Death Certification Anomalies",
        subtitle = "Death certification date appears before the recorded death date",
        x = "Years Before Death",
        y = "Patient"
      ) +
      theme_minimal() +
      theme(
        plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
        plot.subtitle = element_text(hjust = 0.5, size = 11),
        axis.text.y = element_text(face = "bold", size = 9)
      )
    
    p_death_years
    
    ggplotly(p_death_years, tooltip = "text")
  }
}


```

## Negative HbA1c Value Anomaly

```{r anomaly-negative-hba1c-values, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

# HbA1c VALUE numeric deyilsə, numeric-ə çeviririk
observations[, value_num := suppressWarnings(as.numeric(VALUE))]

# Negative raw HbA1c observation-ları
hba1c_raw_negative <- observations[
  DESCRIPTION == "Hemoglobin A1c/Hemoglobin.total in Blood" &
    !is.na(value_num) &
    value_num < 0,
  .(
    PATIENT,
    DATE,
    DESCRIPTION,
    VALUE,
    value_num,
    UNITS
  )
][order(PATIENT, DATE)]

# Ümumi summary: təkrarlarla birlikdə və unique patient sayı
negative_hba1c_count_summary <- hba1c_raw_negative[
  ,
  .(
    total_negative_records = .N,
    unique_patients = uniqueN(PATIENT),
    min_negative_value = min(value_num, na.rm = TRUE),
    max_negative_value = max(value_num, na.rm = TRUE)
  )
]

# Hər xəstədə neçə dəfə negative HbA1c təkrarlanıb
negative_hba1c_by_patient <- hba1c_raw_negative[
  ,
  .(
    negative_record_count = .N,
    min_hba1c = min(value_num, na.rm = TRUE),
    max_hba1c = max(value_num, na.rm = TRUE)
  ),
  by = PATIENT
][order(-negative_record_count, min_hba1c)]

# Nəticələri göstəririk
negative_hba1c_count_summary

negative_hba1c_by_patient

hba1c_raw_negative

# =========================
# Chart 1: Negative HbA1c dəyərlərinin paylanması
# =========================

p_negative_hba1c_values <- ggplot(
  hba1c_raw_negative,
  aes(
    x = value_num,
    text = paste0(
      "Patient ID: ", PATIENT,
      "<br>Date: ", DATE,
      "<br>HbA1c Value: ", value_num,
      "<br>Unit: ", UNITS
    )
  )
) +
  geom_histogram(
    bins = 15,
    color = "white"
  ) +
  geom_vline(
    xintercept = 0,
    linetype = "dashed",
    linewidth = 1
  ) +
  labs(
    title = "Negative HbA1c Value Anomaly",
    subtitle = "Only HbA1c values below 0 are shown",
    x = "HbA1c Value",
    y = "Record Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 11
    )
  )

p_negative_hba1c_values

ggplotly(p_negative_hba1c_values, tooltip = "text")


# =========================
# Chart 2: Patient üzrə negative HbA1c təkrar sayı
# =========================

negative_hba1c_by_patient[, patient_short := paste0(
  substr(PATIENT, 1, 8),
  "..."
)]

top_negative_hba1c_patients <- negative_hba1c_by_patient[
  order(-negative_record_count, min_hba1c)
][1:15]

patients <- as.data.table(patients)

patient_names <- patients[
  ,
  .(
    PATIENT = ID,
    FIRST,
    LAST
  )
]

# Ad-soyadda rəqəm varsa təmizləyirik
patient_names[, FIRST := gsub("[0-9]", "", FIRST)]
patient_names[, LAST := gsub("[0-9]", "", LAST)]

patient_names[, full_name := trimws(paste(FIRST, LAST))]

# Ad boşdursa Patient ID göstərilsin
patient_names[
  is.na(full_name) | full_name == "",
  full_name := paste0("Unknown Patient")
]

negative_hba1c_by_patient_named <- merge(
  negative_hba1c_by_patient,
  patient_names,
  by = "PATIENT",
  all.x = TRUE
)

# Unique label: ad soyad + qısa ID
negative_hba1c_by_patient_named[, patient_label := paste0(
  full_name,
  " (",
  substr(PATIENT, 1, 6),
  "...)"
)]

top_negative_hba1c_patients <- negative_hba1c_by_patient_named[
  order(-negative_record_count, min_hba1c)
][1:15]

p_negative_hba1c_patients <- ggplot(
  top_negative_hba1c_patients,
  aes(
    x = reorder(patient_label, negative_record_count),
    y = negative_record_count,
    fill = negative_record_count,
    text = paste0(
      "Patient Name: ", full_name,
      "<br>Patient ID: ", PATIENT,
      "<br>Negative Record Count: ", negative_record_count,
      "<br>Minimum HbA1c: ", min_hba1c,
      "<br>Maximum HbA1c: ", max_hba1c
    )
  )
) +
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  geom_text(
    aes(label = negative_record_count),
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "Repeated Negative HbA1c Records by Patient",
    subtitle = "Top patients with repeated impossible HbA1c values",
    x = "Patient",
    y = "Negative HbA1c Record Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(
      hjust = 0.5,
      face = "bold",
      size = 18
    ),
    plot.subtitle = element_text(
      hjust = 0.5,
      size = 11
    ),
    axis.text.y = element_text(
      face = "bold",
      size = 8
    ),
    legend.position = "none"
  )

p_negative_hba1c_patients

ggplotly(p_negative_hba1c_patients, tooltip = "text")

```




## Abnormal Encounters After Death

```{r anomaly-after-death-encounters, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(ggplot2)
library(plotly)
library(scales)

patients <- as.data.table(patients)
encounters <- as.data.table(encounters)

# =========================
# 1. Tarixləri hazırlayırıq
# =========================

patients[, death_date := as.IDate(DEATHDATE)]
encounters[, encounter_date := as.IDate(DATE)]

# =========================
# 2. Encounter + patient death date join
# =========================

enc_death <- encounters[
  patients[, .(
    ID,
    death_date,
    BIRTHDATE,
    GENDER,
    RACE,
    ETHNICITY
  )],
  on = c(PATIENT = "ID"),
  nomatch = 0
]

# =========================
# 3. Ölüm tarixindən sonra olan encounter-lər
# =========================

after_death_encounters <- enc_death[
  !is.na(death_date) &
    !is.na(encounter_date) &
    encounter_date > death_date,
  .(
    encounter_id = ID,
    PATIENT,
    encounter_date,
    death_date,
    days_after_death = as.integer(encounter_date - death_date),
    encounter_type = DESCRIPTION,
    reason_description = REASONDESCRIPTION,
    GENDER,
    RACE,
    ETHNICITY
  )
][order(-days_after_death)]

after_death_encounters

# =========================
# 4. Administrative vs abnormal clinical encounter
# =========================

after_death_encounters[, after_death_status := case_when(
  grepl("Death Certification", encounter_type, ignore.case = TRUE) ~
    "Administrative / Death Certification",
  
  TRUE ~
    "Abnormal Clinical Encounter After Death"
)]

# =========================
# 5. Summary: status üzrə
# =========================

after_death_status_summary <- after_death_encounters[
  ,
  .(
    record_count = .N,
    unique_patients = uniqueN(PATIENT),
    min_days_after_death = min(days_after_death, na.rm = TRUE),
    max_days_after_death = max(days_after_death, na.rm = TRUE)
  ),
  by = after_death_status
][order(-record_count)]

after_death_status_summary

# =========================
# 6. Sadəcə abnormal olanlar
# =========================

abnormal_after_death <- after_death_encounters[
  after_death_status == "Abnormal Clinical Encounter After Death"
]

abnormal_after_death

# =========================
# 7. Abnormal encounter type üzrə summary
# =========================

abnormal_after_death_type_summary <- abnormal_after_death[
  ,
  .(
    record_count = .N,
    unique_patients = uniqueN(PATIENT),
    min_days_after_death = min(days_after_death, na.rm = TRUE),
    max_days_after_death = max(days_after_death, na.rm = TRUE)
  ),
  by = encounter_type
][order(-record_count)]

abnormal_after_death_type_summary

# =========================
# 8. Abnormal reason üzrə summary
# =========================

abnormal_after_death_reason_summary <- abnormal_after_death[
  ,
  .(
    record_count = .N,
    unique_patients = uniqueN(PATIENT),
    min_days_after_death = min(days_after_death, na.rm = TRUE),
    max_days_after_death = max(days_after_death, na.rm = TRUE)
  ),
  by = .(encounter_type, reason_description)
][order(-record_count)]

abnormal_after_death_reason_summary

# =========================
# 9. Chart 1: Administrative vs abnormal
# =========================

p_after_death_status <- ggplot(
  after_death_status_summary,
  aes(
    x = after_death_status,
    y = record_count,
    fill = after_death_status,
    text = paste0(
      "Status: ", after_death_status,
      "<br>Record Count: ", scales::comma(record_count),
      "<br>Unique Patients: ", scales::comma(unique_patients),
      "<br>Min Days After Death: ", min_days_after_death,
      "<br>Max Days After Death: ", max_days_after_death
    )
  )
) +
  geom_col(width = 0.65, color = "white") +
  geom_text(
    aes(label = scales::comma(record_count)),
    vjust = -0.3,
    size = 4.5,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "After-Death Encounter Classification",
    subtitle = "Death Certification was separated from abnormal clinical encounters",
    x = "",
    y = "Record Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold", size = 10),
    legend.position = "none"
  )

p_after_death_status

ggplotly(p_after_death_status, tooltip = "text")


# =========================
# 10. Chart 2: Abnormal clinical encounter types
# =========================

if (nrow(abnormal_after_death_type_summary) > 0) {
  
  abnormal_after_death_type_summary[, encounter_type := factor(
    encounter_type,
    levels = rev(encounter_type)
  )]
  
  p_abnormal_after_death_type <- ggplot(
    abnormal_after_death_type_summary,
    aes(
      x = encounter_type,
      y = record_count,
      fill = encounter_type,
      text = paste0(
        "Encounter Type: ", encounter_type,
        "<br>Record Count: ", scales::comma(record_count),
        "<br>Unique Patients: ", scales::comma(unique_patients),
        "<br>Min Days After Death: ", min_days_after_death,
        "<br>Max Days After Death: ", max_days_after_death
      )
    )
  ) +
    geom_col(width = 0.7, color = "white") +
    geom_text(
      aes(label = scales::comma(record_count)),
      hjust = -0.1,
      size = 4,
      fontface = "bold"
    ) +
    coord_flip() +
    scale_y_continuous(
      labels = scales::comma,
      expand = expansion(mult = c(0, 0.25))
    ) +
    labs(
      title = "Abnormal Clinical Encounters After Death",
      subtitle = "Only non-Death-Certification encounters are shown",
      x = "",
      y = "Record Count"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 11),
      axis.text.y = element_text(face = "bold", size = 9),
      legend.position = "none"
    )
  
  p_abnormal_after_death_type
  
  
} else {
  
  print("No abnormal clinical encounters after death were found.")
}
```

## Zero HDL Cholesterol Anomaly

```{r anomaly-zero-hdl-cholesterol, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(ggplot2)
library(plotly)
library(scales)

observations <- as.data.table(observations)
observations[, value_num := suppressWarnings(as.numeric(VALUE))]

hdl_zero <- observations[
  DESCRIPTION == "High Density Lipoprotein Cholesterol" &
    !is.na(value_num) &
    value_num == 0,
  .(
    PATIENT,
    DATE,
    DESCRIPTION,
    VALUE,
    hdl_value = value_num,
    UNITS
  )
][order(PATIENT, DATE)]

hdl_zero_summary <- hdl_zero[
  ,
  .(
    zero_hdl_records = .N,
    unique_patients = uniqueN(PATIENT)
  )
]

hdl_zero_summary
hdl_zero

p_hdl_zero <- ggplot(
  hdl_zero_summary,
  aes(
    x = "HDL = 0",
    y = zero_hdl_records,
    text = paste0(
      "Zero HDL Records: ", scales::comma(zero_hdl_records),
      "<br>Unique Patients: ", scales::comma(unique_patients)
    )
  )
) +
  geom_col(width = 0.5, color = "white") +
  geom_text(
    aes(label = scales::comma(zero_hdl_records)),
    vjust = -0.3,
    size = 5,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Zero HDL Cholesterol Anomaly",
    subtitle = "HDL cholesterol value of 0 is clinically unusual and should be reviewed",
    x = "",
    y = "Record Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11)
  )

p_hdl_zero

ggplotly(p_hdl_zero, tooltip = "text")
```
## Observation and Encounter Year Gap Anomaly

```{r anomaly-observation-encounter-year-gap, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(ggplot2)
library(plotly)
library(scales)

# ===============================
# 1. Date columns hazırlayırıq
# ===============================

obs_dates <- observations[
  ,
  .(
    PATIENT,
    ENCOUNTER,
    obs_date = as.IDate(DATE),
    obs_description = DESCRIPTION,
    VALUE,
    UNITS
  )
]

enc_dates <- encounters[
  ,
  .(
    encounter_id = ID,
    enc_patient = PATIENT,
    encounter_date = as.IDate(DATE),
    encounter_type = DESCRIPTION,
    reason_description = REASONDESCRIPTION
  )
]

# ===============================
# 2. Observation + Encounter join
# ===============================

obs_enc_date_gap <- merge(
  obs_dates,
  enc_dates,
  by.x = "ENCOUNTER",
  by.y = "encounter_id",
  all.x = FALSE
)

# ===============================
# 3. Tarix fərqi: days və years
# ===============================

obs_enc_date_gap[, diff_days := as.integer(obs_date - encounter_date)]

obs_enc_date_gap[, diff_years := round(diff_days / 365.25, 2)]

obs_enc_date_gap[, abs_diff_years := abs(diff_years)]

# Observation encounter-dən əvvəl/sonra
obs_enc_date_gap[, date_direction := case_when(
  diff_years < 0 ~ "Observation before encounter",
  diff_years == 0 ~ "Same year / same date",
  diff_years > 0 ~ "Observation after encounter"
)]

# Year gap category
obs_enc_date_gap[, year_gap_group := case_when(
  abs_diff_years > 1 & abs_diff_years <= 3 ~ "1-3 years",
  abs_diff_years > 3 & abs_diff_years <= 6 ~ "3-6 years",
  abs_diff_years > 6 ~ "6+ years",
  TRUE ~ "Unknown"
)]

obs_enc_date_gap[, year_gap_group := factor(
  year_gap_group,
  levels = c("0-1 years", "1-3 years", "3-6 years", "6+ years")
)]

# ===============================
# 4. Summary table
# ===============================

year_gap_summary <- obs_enc_date_gap[
  ,
  .(
    record_count = .N,
    percent = round(.N / nrow(obs_enc_date_gap) * 100, 2),
    avg_abs_gap_years = round(mean(abs_diff_years, na.rm = TRUE), 2),
    max_abs_gap_years = round(max(abs_diff_years, na.rm = TRUE), 2)
  ),
  by = year_gap_group
][order(year_gap_group)]

year_gap_summary

# Direction üzrə summary
year_gap_direction_summary <- obs_enc_date_gap[
  ,
  .N,
  by = .(year_gap_group, date_direction)
][order(year_gap_group, date_direction)]

year_gap_direction_summary

# ===============================
# 5. Ən uzun fərqlər
# ===============================

longest_year_gap <- obs_enc_date_gap[
  abs_diff_years > 6,
  .(
    PATIENT,
    ENCOUNTER,
    obs_date,
    encounter_date,
    diff_years,
    abs_diff_years,
    date_direction,
    obs_description,
    VALUE,
    UNITS,
    encounter_type,
    reason_description
  )
][order(-abs_diff_years)]

head(longest_year_gap, 30)

# ===============================
# 6. Bar chart
# ===============================

p_year_gap <- ggplot(
  year_gap_summary,
  aes(
    x = year_gap_group,
    y = record_count,
    fill = year_gap_group,
    text = paste0(
      "Year Gap Group: ", year_gap_group,
      "<br>Record Count: ", scales::comma(record_count),
      "<br>Percent: ", percent, "%",
      "<br>Average Gap: ", avg_abs_gap_years, " years",
      "<br>Max Gap: ", max_abs_gap_years, " years"
    )
  )
) +
  geom_col(width = 0.65, color = "white") +
  geom_text(
    aes(label = paste0(scales::comma(record_count), "\n", percent, "%")),
    vjust = -0.3,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Observation and Encounter Date Gap",
    subtitle = "Large year gaps may indicate timeline consistency anomalies",
    x = "Year Gap Group",
    y = "Observation Record Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold", size = 10),
    legend.position = "none"
  )

p_year_gap

ggplotly(p_year_gap, tooltip = "text")

```
## BMI Consistency Anomaly: Height + Weight vs Recorded BMI

```{r anomaly-bmi-consistency-check, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

height_obs <- observations[DESCRIPTION == "Body Height" & !is.na(value_num),
                           .(PATIENT, ENCOUNTER, height = value_num)]
weight_obs <- observations[DESCRIPTION == "Body Weight" & !is.na(value_num),
                           .(PATIENT, ENCOUNTER, weight = value_num)]
bmi_obs    <- observations[DESCRIPTION == "Body Mass Index" & !is.na(value_num),
                           .(PATIENT, ENCOUNTER, bmi_recorded = value_num)]

bmi_check <- merge(height_obs, weight_obs, by = c("PATIENT", "ENCOUNTER")) |>
             merge(bmi_obs,    by = c("PATIENT", "ENCOUNTER"))

bmi_check[, bmi_calc := round(weight / (height / 100)^2, 1)]
bmi_check[, ferq     := abs(bmi_recorded - bmi_calc)]

bmi_incon <- bmi_check[ferq > 2][order(-ferq)]

cat("=== 1) BMI Tutarsızlığı (fərq > 2) ===\n")
cat("Tutarsız ölçü sayı:", nrow(bmi_incon), "\n")
if (nrow(bmi_incon) > 0) {
  cat("Maksimum fərq:", max(bmi_incon$ferq), "\n")
  print(head(bmi_incon[, .(PATIENT, height, weight, bmi_recorded, bmi_calc, ferq)], 5))
}

# =========================
# Chart: BMI tutarsızlığı summary
# =========================

library(ggplot2)
library(plotly)
library(scales)

bmi_check[, ferq_group := fifelse(
  ferq <= 2, "Consistent / acceptable",
  fifelse(
    ferq > 2 & ferq <= 5, "Mismatch: 2-5",
    fifelse(
      ferq > 5 & ferq <= 10, "Mismatch: 5-10",
      "Mismatch: 10+"
    )
  )
)]

bmi_check[, ferq_group := factor(
  ferq_group,
  levels = c(
    "Consistent / acceptable",
    "Mismatch: 2-5",
    "Mismatch: 5-10",
    "Mismatch: 10+"
  )
)]

bmi_chart_data <- bmi_check[
  ,
  .(
    record_count = .N,
    avg_difference = round(mean(ferq, na.rm = TRUE), 2),
    max_difference = round(max(ferq, na.rm = TRUE), 2)
  ),
  by = ferq_group
][order(ferq_group)]

p_bmi_mismatch <- ggplot(
  bmi_chart_data,
  aes(
    x = ferq_group,
    y = record_count,
    fill = ferq_group,
    text = paste0(
      "Group: ", ferq_group,
      "<br>Record Count: ", scales::comma(record_count),
      "<br>Average Difference: ", avg_difference,
      "<br>Max Difference: ", max_difference
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = scales::comma(record_count)),
    vjust = -0.3,
    size = 4,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "BMI Consistency Check",
    subtitle = "Recorded BMI compared with BMI calculated from height and weight",
    x = "BMI Difference Group",
    y = "Record Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold", size = 10),
    legend.position = "none"
  )

p_bmi_mismatch

ggplotly(p_bmi_mismatch, tooltip = "text")
```
## Duplicate Passport Anomaly

```{r anomaly-duplicate-passport, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(ggplot2)
library(plotly)
library(scales)

patients <- as.data.table(patients)

# =========================
# 1. Passport values clean
# =========================

patients[, passport_clean := trimws(PASSPORT)]

# Placeholder / invalid passport values çıxarılır
passport_valid <- patients[
  !is.na(passport_clean) &
    passport_clean != "" &
    tolower(passport_clean) != "false",
  .(
    ID,
    FIRST,
    LAST,
    GENDER,
    BIRTHDATE,
    DEATHDATE,
    PASSPORT = passport_clean
  )
]

# Ad soyad təmizlənir
passport_valid[, FIRST := gsub("[0-9]", "", FIRST)]
passport_valid[, LAST := gsub("[0-9]", "", LAST)]
passport_valid[, patient_name := trimws(paste(FIRST, LAST))]

passport_valid[
  is.na(patient_name) | patient_name == "",
  patient_name := "Unknown Patient"
]

# =========================
# 2. Duplicate passport summary
# =========================

duplicate_passport_summary <- passport_valid[
  ,
  .(
    patient_count = uniqueN(ID),
    patient_names = paste(unique(patient_name), collapse = "; "),
    patient_ids = paste(unique(ID), collapse = "; ")
  ),
  by = PASSPORT
][
  patient_count > 1
][order(-patient_count)]

duplicate_passport_summary

# Ümumi anomaly summary
duplicate_passport_count_summary <- duplicate_passport_summary[
  ,
  .(
    duplicate_passport_codes = .N,
    affected_patients = sum(patient_count),
    max_patients_same_passport = max(patient_count, na.rm = TRUE)
  )
]



# =========================
# 3. Detailed duplicate records
# =========================

duplicate_passport_records <- passport_valid[
  PASSPORT %in% duplicate_passport_summary$PASSPORT
][order(PASSPORT, patient_name)]

duplicate_passport_records

# =========================
# 4. Top duplicate passport chart data
# =========================

top_duplicate_passports <- duplicate_passport_summary[
  order(-patient_count)
][1:min(15, .N)]

top_duplicate_passports[, passport_label := paste0(
  substr(PASSPORT, 1, 10),
  "..."
)]

top_duplicate_passports[, passport_label := factor(
  passport_label,
  levels = rev(passport_label)
)]

# =========================
# 5. Chart
# =========================

if (nrow(top_duplicate_passports) > 0) {
  
  p_duplicate_passport <- ggplot(
    top_duplicate_passports,
    aes(
      x = passport_label,
      y = patient_count,
      fill = patient_count,
      text = paste0(
        "Passport: ", PASSPORT,
        "<br>Patient Count: ", patient_count,
        "<br>Patient Names: ", patient_names
      )
    )
  ) +
    geom_col(width = 0.7, color = "white") +
    geom_text(
      aes(label = patient_count),
      hjust = -0.1,
      size = 4,
      fontface = "bold"
    ) +
    coord_flip() +
    scale_y_continuous(
      labels = scales::comma,
      expand = expansion(mult = c(0, 0.25))
    ) +
    labs(
      title = "Duplicate Passport Code Anomaly",
      subtitle = "Same passport code appearing for more than one patient",
      x = "Passport Code",
      y = "Patient Count"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 11),
      axis.text.y = element_text(face = "bold", size = 9),
      legend.position = "none"
    )
  duplicate_passport_count_summary
  p_duplicate_passport
  
  
} else {
  print("No duplicate valid passport codes found.")
}
```
## Shared Passport by Patients Anomaly

```{r anomaly-shared-passport-patients, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(ggplot2)
library(plotly)
library(scales)

patients <- as.data.table(patients)

# =========================
# 1. Passport clean
# =========================

patients[, passport_clean := trimws(PASSPORT)]

passport_valid <- patients[
  !is.na(passport_clean) &
    passport_clean != "" &
    tolower(passport_clean) != "false",
  .(
    ID,
    FIRST,
    LAST,
    GENDER,
    BIRTHDATE,
    PASSPORT = passport_clean
  )
]

# Ad-soyad təmizliyi
passport_valid[, FIRST := gsub("[0-9]", "", FIRST)]
passport_valid[, LAST := gsub("[0-9]", "", LAST)]
passport_valid[, patient_name := trimws(paste(FIRST, LAST))]

passport_valid[
  is.na(patient_name) | patient_name == "",
  patient_name := "Unknown Patient"
]

# =========================
# 2. Eyni passport neçə patient-də var?
# =========================

passport_groups <- passport_valid[
  ,
  .(
    patient_count = uniqueN(ID),
    patient_names = paste(unique(patient_name), collapse = " / "),
    patient_ids = paste(unique(ID), collapse = " / ")
  ),
  by = PASSPORT
][
  patient_count %in% c(2, 3)
][order(-patient_count)]

passport_groups

# =========================
# 3. Chart label: passport yox, patient names göstərilir
# =========================

passport_groups[, group_label := paste0(
  patient_names,
  " (",
  patient_count,
  " patients)"
)]

# Əgər çox uzundursa, qısalt
passport_groups[, group_label_short := ifelse(
  nchar(group_label) > 55,
  paste0(substr(group_label, 1, 55), "..."),
  group_label
)]

# Top 15 group göstəririk ki chart chaotic olmasın
top_passport_groups <- passport_groups[
  order(-patient_count)
][1:min(15, .N)]

top_passport_groups[, group_label_short := factor(
  group_label_short,
  levels = rev(group_label_short)
)]

# =========================
# 4. Chart: patient names sharing same passport
# =========================

p_shared_passport_patients <- ggplot(
  top_passport_groups,
  aes(
    x = group_label_short,
    y = patient_count,
    fill = patient_count,
    text = paste0(
      "Shared Passport: ", PASSPORT,
      "<br>Patient Count: ", patient_count,
      "<br>Patients: ", patient_names,
      "<br>Patient IDs: ", patient_ids
    )
  )
) +
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  geom_text(
    aes(label = patient_count),
    hjust = -0.2,
    size = 4.5,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    breaks = c(0, 1, 2, 3),
    limits = c(0, 3.5),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Shared Passport Anomaly by Patient Names",
    subtitle = "Patient groups where the same passport code appears for 2 or 3 patients",
    x = "Patients Sharing the Same Passport",
    y = "Patient Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 8),
    legend.position = "none"
  )

p_shared_passport_patients

ggplotly(p_shared_passport_patients, tooltip = "text")
```

## Duplicate SSN Anomaly

```{r anomaly-duplicate-ssn, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(ggplot2)
library(plotly)
library(scales)

patients <- as.data.table(patients)

# =========================
# 1. SSN clean
# =========================

patients[, ssn_clean := trimws(SSN)]

# Placeholder / invalid SSN values çıxarılır
ssn_valid <- patients[
  !is.na(ssn_clean) &
    ssn_clean != "" &
    tolower(ssn_clean) != "false",
  .(
    ID,
    FIRST,
    LAST,
    GENDER,
    BIRTHDATE,
    SSN = ssn_clean
  )
]

# Ad-soyad təmizliyi
ssn_valid[, FIRST := gsub("[0-9]", "", FIRST)]
ssn_valid[, LAST := gsub("[0-9]", "", LAST)]
ssn_valid[, patient_name := trimws(paste(FIRST, LAST))]

ssn_valid[
  is.na(patient_name) | patient_name == "",
  patient_name := "Unknown Patient"
]

# =========================
# 2. Hər SSN neçə patient-də görünür?
# =========================

ssn_count <- ssn_valid[
  ,
  .(
    patient_count = uniqueN(ID),
    patient_names = paste(unique(patient_name), collapse = "; "),
    patient_ids = paste(unique(ID), collapse = "; ")
  ),
  by = SSN
]

# Təkrarlanan SSN-lər
duplicate_ssn <- ssn_count[
  patient_count > 1
][order(-patient_count)]

duplicate_ssn

# =========================
# 3. Summary
# =========================

duplicate_ssn_summary <- duplicate_ssn[
  ,
  .(
    repeated_ssn_codes = .N,
    affected_patients = sum(patient_count),
    max_patients_same_ssn = max(patient_count, na.rm = TRUE)
  )
]

duplicate_ssn_summary

# =========================
# 4. Duplicate / Triple / Multiple group
# =========================

duplicate_ssn[, ssn_anomaly_group := case_when(
  patient_count == 2 ~ "Duplicate SSN",
  patient_count == 3 ~ "Triple SSN",
  patient_count > 3 ~ "Multiple SSN",
  TRUE ~ "Other"
)]

ssn_anomaly_summary <- duplicate_ssn[
  ,
  .(
    ssn_code_count = .N,
    affected_patient_count = sum(patient_count)
  ),
  by = ssn_anomaly_group
]

# Əgər hansısa group yoxdursa, 0 kimi saxlayırıq
all_ssn_groups <- data.table(
  ssn_anomaly_group = c(
    "Duplicate SSN",
    "Triple SSN",
    "Multiple SSN"
  )
)

ssn_anomaly_summary <- merge(
  all_ssn_groups,
  ssn_anomaly_summary,
  by = "ssn_anomaly_group",
  all.x = TRUE
)

ssn_anomaly_summary[is.na(ssn_code_count), ssn_code_count := 0]
ssn_anomaly_summary[is.na(affected_patient_count), affected_patient_count := 0]

ssn_anomaly_summary[, ssn_anomaly_group := factor(
  ssn_anomaly_group,
  levels = c("Duplicate SSN", "Triple SSN", "Multiple SSN")
)]

ssn_anomaly_summary

# =========================
# 5. Chart: SSN anomaly summary
# =========================

p_ssn_anomaly <- ggplot(
  ssn_anomaly_summary,
  aes(
    x = ssn_anomaly_group,
    y = ssn_code_count,
    fill = ssn_anomaly_group,
    text = paste0(
      "Anomaly Group: ", ssn_anomaly_group,
      "<br>SSN Code Count: ", scales::comma(ssn_code_count),
      "<br>Affected Patient Count: ", scales::comma(affected_patient_count)
    )
  )
) +
  geom_col(
    width = 0.65,
    color = "white"
  ) +
  geom_text(
    aes(label = scales::comma(ssn_code_count)),
    vjust = -0.3,
    size = 5,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "Repeated SSN Anomaly",
    subtitle = "Same SSN number appearing for more than one patient",
    x = "",
    y = "SSN Code Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold", size = 11),
    legend.position = "none"
  )

p_ssn_anomaly

ggplotly(p_ssn_anomaly, tooltip = "text")
```

# Hypothesis Analysis

## Cancer/Tumor Patients and Average Glucose Hypothesis

```{r hypothesis-cancer-glucose-hba1c, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(ggplot2)
library(plotly)
library(scales)

master <- as.data.table(master)
encounters <- as.data.table(encounters)
observations <- as.data.table(observations)

observations[, value_num := suppressWarnings(as.numeric(VALUE))]

encounters[, reason_group := case_when(
  
  grepl(
    "malignant|cancer|tumor|neoplasm|carcinoma",
    REASONDESCRIPTION,
    ignore.case = TRUE
  ) ~ "Cancer / Tumor",
  
  grepl(
    "diabetes|hypertension|asthma|copd|emphysema|chronic",
    REASONDESCRIPTION,
    ignore.case = TRUE
  ) ~ "Chronic Disease",
  
  grepl(
    "sinusitis|bronchitis|pharyngitis|otitis|sore throat|viral|bacterial|strep",
    REASONDESCRIPTION,
    ignore.case = TRUE
  ) ~ "Acute Infection / ENT",
  
  grepl(
    "fracture|injury|trauma|concussion|sprain|burn",
    REASONDESCRIPTION,
    ignore.case = TRUE
  ) ~ "Injury / Trauma",
  
  grepl(
    "pregnancy|prenatal|childbirth|miscarriage",
    REASONDESCRIPTION,
    ignore.case = TRUE
  ) ~ "Pregnancy Related",
  
  is.na(REASONDESCRIPTION) | trimws(REASONDESCRIPTION) == "" ~ "No Reason Recorded",
  
  TRUE ~ "Other"
)]

patient_reason_group <- encounters[
  reason_group != "No Reason Recorded",
  .N,
  by = .(PATIENT, reason_group)
][
  order(PATIENT, -N)
][
  ,
  .SD[1],
  by = PATIENT
]

setnames(patient_reason_group, "N", "reason_group_count")

old_reason_cols <- intersect(
  names(master),
  c("reason_group", "reason_group_count")
)

if (length(old_reason_cols) > 0) {
  master[, (old_reason_cols) := NULL]
}

master <- merge(
  master,
  patient_reason_group,
  by.x = "ID",
  by.y = "PATIENT",
  all.x = TRUE
)

master[is.na(reason_group), reason_group := "No Reason Recorded"]
master[is.na(reason_group_count), reason_group_count := 0]

glucose_data <- observations[
  DESCRIPTION == "Glucose" &
    !is.na(value_num),
  .(
    PATIENT,
    DATE,
    glucose = value_num
  )
][
  order(PATIENT, DATE)
][
  ,
  .SD[.N],
  by = PATIENT
]

if ("glucose" %in% names(master)) {
  master[, glucose := NULL]
}

master <- merge(
  master,
  glucose_data[, .(PATIENT, glucose)],
  by.x = "ID",
  by.y = "PATIENT",
  all.x = TRUE
)

master[, glucose := as.numeric(glucose)]
comparison_groups <- c(
  "Cancer / Tumor",
  "Chronic Disease",
  "Acute Infection / ENT",
  "Injury / Trauma",
  "Pregnancy Related"
)

cancer_glucose_summary <- master[
  reason_group %in% comparison_groups &
    !is.na(glucose),
  .(
    patient_count = .N,
    avg_glucose = round(mean(glucose, na.rm = TRUE), 1),
    median_glucose = round(median(glucose, na.rm = TRUE), 1),
    min_glucose = min(glucose, na.rm = TRUE),
    max_glucose = max(glucose, na.rm = TRUE)
  ),
  by = reason_group
][order(-avg_glucose)]

cancer_glucose_summary

cancer_rank <- cancer_glucose_summary[
  ,
  .(
    reason_group,
    avg_glucose,
    glucose_rank = frank(-avg_glucose, ties.method = "min")
  )
][order(glucose_rank)]

cancer_rank

p_cancer_glucose <- ggplot(
  cancer_glucose_summary,
  aes(
    x = reorder(reason_group, avg_glucose),
    y = avg_glucose,
    fill = reason_group,
    text = paste0(
      "Reason Group: ", reason_group,
      "<br>Patient Count: ", scales::comma(patient_count),
      "<br>Average Glucose: ", avg_glucose,
      "<br>Median Glucose: ", median_glucose,
      "<br>Min Glucose: ", min_glucose,
      "<br>Max Glucose: ", max_glucose
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = paste0(avg_glucose, "\n(n=", scales::comma(patient_count), ")")),
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    expand = expansion(mult = c(0, 0.22))
  ) +
  labs(
    title = "Average Glucose by Reason Group",
    subtitle = "Testing whether Cancer/Tumor patients have the highest average glucose",
    x = "",
    y = "Average Glucose"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 9),
    legend.position = "none"
  )
p_cancer_glucose
ggplotly(p_cancer_glucose, tooltip = "text")

```


## Hidden Clinical Risk in Outpatient Encounters Hypothesis

```{r hypothesis-misleading-normal-visit-risk, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(ggplot2)
library(plotly)
library(scales)

master <- as.data.table(master)
encounters <- as.data.table(encounters)
observations <- as.data.table(observations)


observations[, value_num := suppressWarnings(as.numeric(VALUE))]

risk_obs <- observations[
  DESCRIPTION %in% c(
    "Body Mass Index",
    "Systolic Blood Pressure",
    "Diastolic Blood Pressure",
    "Glucose",
    "Hemoglobin A1c/Hemoglobin.total in Blood",
    "Total Cholesterol",
    "Triglycerides"
  ) &
    !is.na(value_num),
  .(
    PATIENT,
    DATE,
    DESCRIPTION,
    value_num
  )
][
  order(PATIENT, DESCRIPTION, DATE)
][
  ,
  .SD[.N],
  by = .(PATIENT, DESCRIPTION)
]

risk_wide <- dcast(
  risk_obs,
  PATIENT ~ DESCRIPTION,
  value.var = "value_num"
)

setDT(risk_wide)

setnames(
  risk_wide,
  old = c(
    "Body Mass Index",
    "Systolic Blood Pressure",
    "Diastolic Blood Pressure",
    "Glucose",
    "Hemoglobin A1c/Hemoglobin.total in Blood",
    "Total Cholesterol",
    "Triglycerides"
  ),
  new = c(
    "bmi",
    "sbp",
    "dbp",
    "glucose",
    "hba1c",
    "cholesterol",
    "triglycerides"
  ),
  skip_absent = TRUE
)

# Əgər hansısa column yoxdursa, NA yaradılır
needed_cols <- c(
  "bmi",
  "sbp",
  "dbp",
  "glucose",
  "hba1c",
  "cholesterol",
  "triglycerides"
)

for (col in needed_cols) {
  if (!col %in% names(risk_wide)) {
    risk_wide[, (col) := NA_real_]
  }
}

risk_wide[, severe_obesity := ifelse(!is.na(bmi) & bmi >= 40, 1, 0)]

risk_wide[, crisis_bp := ifelse(
  (!is.na(sbp) & sbp >= 180) |
    (!is.na(dbp) & dbp >= 120),
  1,
  0
)]

risk_wide[, high_glucose := ifelse(
  !is.na(glucose) & glucose >= 200,
  1,
  0
)]

risk_wide[, high_hba1c := ifelse(
  !is.na(hba1c) & hba1c > 0 & hba1c >= 9,
  1,
  0
)]

risk_wide[, high_cholesterol := ifelse(
  !is.na(cholesterol) & cholesterol >= 240,
  1,
  0
)]

risk_wide[, high_triglycerides := ifelse(
  !is.na(triglycerides) & triglycerides >= 200,
  1,
  0
)]

risk_wide[, hidden_risk_count :=
            severe_obesity +
            crisis_bp +
            high_glucose +
            high_hba1c +
            high_cholesterol +
            high_triglycerides]

risk_wide[, hidden_risk_flag := ifelse(
  hidden_risk_count >= 1,
  1,
  0
)]

patient_encounter_type <- encounters[
  ,
  .N,
  by = .(PATIENT, DESCRIPTION)
][
  order(PATIENT, -N)
][
  ,
  .SD[1],
  by = PATIENT
]

setnames(patient_encounter_type, "DESCRIPTION", "dominant_encounter_type")
setnames(patient_encounter_type, "N", "encounter_type_count")

normal_visit_risk <- merge(
  patient_encounter_type,
  risk_wide,
  by = "PATIENT",
  all.x = TRUE
)

normal_visit_risk[is.na(hidden_risk_count), hidden_risk_count := 0]
normal_visit_risk[is.na(hidden_risk_flag), hidden_risk_flag := 0]

normal_visit_risk[, visit_group := ifelse(
  grepl("Outpatient", dominant_encounter_type, ignore.case = TRUE),
  "Outpatient Encounter",
  "Other Encounter Types"
)]

# Summary
misleading_summary <- normal_visit_risk[
  ,
  .(
    patient_count = .N,
    hidden_risk_patients = sum(hidden_risk_flag, na.rm = TRUE),
    hidden_risk_percent = round(mean(hidden_risk_flag, na.rm = TRUE) * 100, 1),
    avg_hidden_risk_count = round(mean(hidden_risk_count, na.rm = TRUE), 2)
  ),
  by = visit_group
][order(-hidden_risk_percent)]

misleading_summary


outpatient_risk_detail <- normal_visit_risk[
  visit_group == "Outpatient Encounter",
  .(
    severe_obesity_percent = round(mean(severe_obesity, na.rm = TRUE) * 100, 1),
    crisis_bp_percent = round(mean(crisis_bp, na.rm = TRUE) * 100, 1),
    high_glucose_percent = round(mean(high_glucose, na.rm = TRUE) * 100, 1),
    high_hba1c_percent = round(mean(high_hba1c, na.rm = TRUE) * 100, 1),
    high_cholesterol_percent = round(mean(high_cholesterol, na.rm = TRUE) * 100, 1),
    high_triglycerides_percent = round(mean(high_triglycerides, na.rm = TRUE) * 100, 1)
  )
]

outpatient_risk_detail

# Long format for chart
outpatient_risk_long <- melt(
  outpatient_risk_detail,
  variable.name = "risk_type",
  value.name = "risk_percent"
)

setDT(outpatient_risk_long)

outpatient_risk_long[, risk_type := case_when(
  risk_type == "severe_obesity_percent" ~ "Severe obesity",
  risk_type == "crisis_bp_percent" ~ "Crisis-level BP",
  risk_type == "high_glucose_percent" ~ "High glucose",
  risk_type == "high_hba1c_percent" ~ "High HbA1c",
  risk_type == "high_cholesterol_percent" ~ "High cholesterol",
  risk_type == "high_triglycerides_percent" ~ "High triglycerides",
  TRUE ~ risk_type
)]

outpatient_risk_long <- outpatient_risk_long[order(-risk_percent)]

outpatient_risk_long

p_misleading_visit <- ggplot(
  misleading_summary,
  aes(
    x = visit_group,
    y = hidden_risk_percent,
    fill = visit_group,
    text = paste0(
      "Visit Group: ", visit_group,
      "<br>Patient Count: ", scales::comma(patient_count),
      "<br>Hidden Risk Patients: ", scales::comma(hidden_risk_patients),
      "<br>Hidden Risk Percent: ", hidden_risk_percent, "%",
      "<br>Average Hidden Risk Count: ", avg_hidden_risk_count
    )
  )
) +
  geom_col(width = 0.65, color = "white") +
  geom_text(
    aes(label = paste0(hidden_risk_percent, "%")),
    vjust = -0.3,
    size = 4.5,
    fontface = "bold"
  ) +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.2))
  ) +
  labs(
    title = "Hidden Clinical Risk in Routine-Looking Visits",
    subtitle = "Outpatient encounters compared with other encounter types",
    x = "",
    y = "Patients With At Least One Hidden Risk Signal"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold", size = 10),
    legend.position = "none"
  )

p_misleading_visit

ggplotly(p_misleading_visit, tooltip = "text")


p_outpatient_risk_types <- ggplot(
  outpatient_risk_long,
  aes(
    x = reorder(risk_type, risk_percent),
    y = risk_percent,
    fill = risk_type,
    text = paste0(
      "Risk Type: ", risk_type,
      "<br>Outpatient Risk Percent: ", risk_percent, "%"
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = paste0(risk_percent, "%")),
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = function(x) paste0(x, "%"),
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "Which Hidden Risks Appear in Outpatient Encounters?",
    subtitle = "Risk signal breakdown among patients whose dominant encounter type is outpatient",
    x = "",
    y = "Outpatient Patient Percentage"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 9),
    legend.position = "none"
  )

p_outpatient_risk_types

ggplotly(p_outpatient_risk_types, tooltip = "text")

```



# Forecast Analysis

# forecast all clinical measurements 2018

```{r forecast-all-clinical-measurements-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(plotly)
library(scales)

observations <- as.data.table(observations)

# =========================
# 1. Date and numeric preparation
# =========================

observations[, obs_date := as.IDate(DATE)]
observations[, obs_year := year(obs_date)]
observations[, value_num := suppressWarnings(as.numeric(VALUE))]

# =========================
# 2. All numeric clinical measurement records
# Burada specific measurement seçmirik.
# Bütün numeric observation measurement-ləri sayılır.
# =========================

measurement_yearly <- observations[
  !is.na(obs_year) &
    !is.na(value_num),
  .(
    clinical_measurement_count = .N
  ),
  by = obs_year
][order(obs_year)]

measurement_yearly

# =========================
# 3. Mark complete / incomplete years
# =========================

measurement_yearly[, year_status := ifelse(
  obs_year == 2017,
  "Incomplete year excluded from model",
  "Complete year used for model"
)]
# =========================
# 4. Training data: 2011-2016 only
# 2010 early ramp-up year excluded
# 2017 incomplete year excluded
# =========================

measurement_train <- measurement_yearly[
  obs_year >= 2011 &
    obs_year <= 2016
]

measurement_train[, year_index := obs_year - min(obs_year) + 1]

measurement_train

# =========================
# 5. FIX: Linear regression model (NOT polynomial)
# Polynomial 6 data nöqtəsində overfitting edir — R² süni yüksəlir.
# Sadə xətti reqressiya + LOOCV istifadə edilir.
# =========================

measurement_lm_model <- lm(
  clinical_measurement_count ~ year_index,
  data = measurement_train
)

summary(measurement_lm_model)

measurement_r2 <- round(summary(measurement_lm_model)$r.squared, 3)
measurement_adj_r2 <- round(summary(measurement_lm_model)$adj.r.squared, 3)

# LOOCV validation
library(caret)
ctrl_loocv <- trainControl(method = "LOOCV")
cv_meas <- suppressWarnings(train(
  clinical_measurement_count ~ year_index,
  data      = as.data.frame(measurement_train),
  method    = "lm",
  trControl = ctrl_loocv
))
measurement_loocv_r2   <- round(cv_meas$results$Rsquared, 3)
measurement_loocv_rmse <- round(cv_meas$results$RMSE, 0)

cat("Train R²:    ", measurement_r2, "\n")
cat("LOOCV R²:    ", measurement_loocv_r2, "(reliable estimate)\n")
cat("LOOCV RMSE:  ", measurement_loocv_rmse, "\n")

measurement_r2
measurement_adj_r2

# =========================
# 6. Forecast 2018
# =========================

forecast_2018 <- data.table(
  obs_year = 2018
)

forecast_2018[, year_index := obs_year - min(measurement_train$obs_year) + 1]

pred_2018 <- predict(
  measurement_lm_model,
  newdata = forecast_2018,
  interval = "prediction"
)

forecast_2018[, forecast_count := round(as.numeric(pred_2018[, "fit"]), 0)]
forecast_2018[, lower_95 := round(as.numeric(pred_2018[, "lwr"]), 0)]
forecast_2018[, upper_95 := round(as.numeric(pred_2018[, "upr"]), 0)]

forecast_2018[forecast_count < 0, forecast_count := 0]
forecast_2018[lower_95 < 0, lower_95 := 0]
forecast_2018[upper_95 < 0, upper_95 := 0]

forecast_2018

# =========================
# 7. Fitted values for 2010-2016
# =========================

measurement_train[, fitted_count := round(
  predict(measurement_lm_model, newdata = measurement_train),
  0
)]

# =========================
# 8. Plot data
# =========================

actual_plot <- measurement_yearly[
  obs_year >= 2010 &
    obs_year <= 2017,
  .(
    obs_year,
    clinical_measurement_count,
    year_status
  )
]

forecast_plot <- forecast_2018[
  ,
  .(
    obs_year,
    forecast_count,
    lower_95,
    upper_95
  )
]

# =========================
# 9. Chart: All clinical measurement forecast
# =========================

p_all_measurement_2018 <- ggplot() +
  
  # Actual complete years
  geom_line(
    data = actual_plot[
      obs_year <= 2016
    ],
    aes(
      x = obs_year,
      y = clinical_measurement_count
    ),
    linewidth = 1.2
  ) +
  
  geom_point(
    data = actual_plot[
      obs_year <= 2016
    ],
    aes(
      x = obs_year,
      y = clinical_measurement_count,
      text = paste0(
        "Year: ", obs_year,
        "<br>All Numeric Measurement Count: ",
        scales::comma(clinical_measurement_count),
        "<br>Status: Complete year used for model"
      )
    ),
    size = 3
  ) +
  
  geom_text(
    data = actual_plot[
      obs_year <= 2016
    ],
    aes(
      x = obs_year,
      y = clinical_measurement_count,
      label = scales::comma(clinical_measurement_count)
    ),
    vjust = -0.8,
    size = 3.5,
    fontface = "bold"
  ) +
  
  # 2017 incomplete point
  geom_point(
    data = actual_plot[
      obs_year == 2017
    ],
    aes(
      x = obs_year,
      y = clinical_measurement_count,
      text = paste0(
        "Year: ", obs_year,
        "<br>All Numeric Measurement Count: ",
        scales::comma(clinical_measurement_count),
        "<br>Status: Incomplete year excluded from model"
      )
    ),
    size = 4,
    shape = 17
  ) +
  
  geom_text(
    data = actual_plot[
      obs_year == 2017
    ],
    aes(
      x = obs_year,
      y = clinical_measurement_count,
      label = paste0(scales::comma(clinical_measurement_count), "\nIncomplete")
    ),
    vjust = 1.4,
    size = 3.8,
    fontface = "bold"
  ) +
  
  # Forecast interval
  geom_errorbar(
    data = forecast_plot,
    aes(
      x = obs_year,
      ymin = lower_95,
      ymax = upper_95
    ),
    width = 0.15,
    linewidth = 1
  ) +
  
  # Forecast point
  geom_point(
    data = forecast_plot,
    aes(
      x = obs_year,
      y = forecast_count,
      text = paste0(
        "Forecast Year: ", obs_year,
        "<br>Forecast Count: ", scales::comma(forecast_count),
        "<br>95% Prediction Interval: ",
        scales::comma(lower_95), " - ",
        scales::comma(upper_95)
      )
    ),
    size = 4
  ) +
  
  geom_text(
    data = forecast_plot,
    aes(
      x = obs_year,
      y = forecast_count,
      label = scales::comma(forecast_count)
    ),
    vjust = -1,
    size = 4.2,
    fontface = "bold"
  ) +
  
  # Connect 2016 to forecast with dashed line
  geom_segment(
    data = data.table(
      x = 2016,
      xend = 2018,
      y = measurement_train[obs_year == 2016, clinical_measurement_count],
      yend = forecast_2018$forecast_count
    ),
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    linetype = "dashed",
    linewidth = 1
  ) +
  
  scale_x_continuous(
    breaks = 2010:2018
  ) +
  
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.05, 0.18))
  ) +
  
  labs(
    title = "2018 Overall Clinical Measurement Count Forecast",
    subtitle = paste0(
      "All numeric observation measurements included; 2017 excluded as incomplete | R² = ",
      measurement_r2
    ),
    x = "Year",
    y = "All Clinical Measurement Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold")
  )

p_all_measurement_2018

ggplotly(p_all_measurement_2018, tooltip = "text")
overall_clinical_measurement_forecast_summary <- data.table(
  model = "Polynomial Regression",
  target = "All numeric clinical measurement records",
  training_period = "2011-2016 complete years",
  excluded_years = "2010 ramp-up year; 2017 incomplete year",
  forecast_year = 2018,
  forecast_count = forecast_2018$forecast_count,
  lower_95 = forecast_2018$lower_95,
  upper_95 = forecast_2018$upper_95,
  r_squared = measurement_r2,
  adjusted_r_squared = measurement_adj_r2
)
```

## Hospital Encounter Forecast
```{r forecast-hospital-encounter-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(plotly)
library(scales)

encounters <- as.data.table(encounters)

# =========================
# 1. Date preparation
# =========================

encounters[, encounter_date := as.IDate(DATE)]
encounters[, encounter_year := year(encounter_date)]

# =========================
# 2. Yearly hospital encounter count
# =========================

encounter_yearly <- encounters[
  !is.na(encounter_year),
  .(
    encounter_count = .N
  ),
  by = encounter_year
][order(encounter_year)]

encounter_yearly

# =========================
# 3. Mark complete / incomplete years
# =========================

encounter_yearly[, year_status := ifelse(
  encounter_year == 2017,
  "Incomplete year excluded from model",
  "Complete year used for model"
)]

# =========================
# 4. Training data: 2010-2016 only
# =========================

encounter_train <- encounter_yearly[
  encounter_year >= 2010 &
    encounter_year <= 2016
]

encounter_train[, year_index := encounter_year - min(encounter_year) + 1]

encounter_train

# =========================
# 5. Linear regression model
# =========================

encounter_lm_model <- lm(
  encounter_count ~ year_index,
  data = encounter_train
)

summary(encounter_lm_model)

# R-squared
encounter_r2 <- round(summary(encounter_lm_model)$r.squared, 3)

encounter_r2

# =========================
# 6. Forecast 2018
# =========================

forecast_2018 <- data.table(
  encounter_year = 2018
)

forecast_2018[, year_index := encounter_year - min(encounter_train$encounter_year) + 1]

pred_2018 <- predict(
  encounter_lm_model,
  newdata = forecast_2018,
  interval = "prediction"
)

forecast_2018[, forecast_count := round(as.numeric(pred_2018[, "fit"]), 0)]
forecast_2018[, lower_95 := round(as.numeric(pred_2018[, "lwr"]), 0)]
forecast_2018[, upper_95 := round(as.numeric(pred_2018[, "upr"]), 0)]

forecast_2018[forecast_count < 0, forecast_count := 0]
forecast_2018[lower_95 < 0, lower_95 := 0]
forecast_2018[upper_95 < 0, upper_95 := 0]

forecast_2018

# =========================
# 7. Fitted values for 2010-2016
# =========================

encounter_train[, fitted_count := round(
  predict(encounter_lm_model, newdata = encounter_train),
  0
)]

# =========================
# 8. Plot data
# =========================

actual_plot <- encounter_yearly[
  encounter_year >= 2010 &
    encounter_year <= 2017,
  .(
    encounter_year,
    encounter_count,
    year_status
  )
]

forecast_plot <- forecast_2018[
  ,
  .(
    encounter_year,
    forecast_count,
    lower_95,
    upper_95
  )
]

# =========================
# 9. Chart: Corrected 2018 encounter forecast
# =========================

p_encounter_2018_fixed <- ggplot() +
  
  # Actual complete years: 2010-2016
  geom_line(
    data = actual_plot[
      encounter_year <= 2016
    ],
    aes(
      x = encounter_year,
      y = encounter_count
    ),
    linewidth = 1.2
  ) +
  
  geom_point(
    data = actual_plot[
      encounter_year <= 2016
    ],
    aes(
      x = encounter_year,
      y = encounter_count,
      text = paste0(
        "Year: ", encounter_year,
        "<br>Encounter Count: ",
        scales::comma(encounter_count),
        "<br>Status: Complete year used for model"
      )
    ),
    size = 3
  ) +
  
  geom_text(
    data = actual_plot[
      encounter_year <= 2016
    ],
    aes(
      x = encounter_year,
      y = encounter_count,
      label = scales::comma(encounter_count)
    ),
    vjust = -0.8,
    size = 3.5,
    fontface = "bold"
  ) +
  
  # 2017 incomplete point
  geom_point(
    data = actual_plot[
      encounter_year == 2017
    ],
    aes(
      x = encounter_year,
      y = encounter_count,
      text = paste0(
        "Year: ", encounter_year,
        "<br>Encounter Count: ",
        scales::comma(encounter_count),
        "<br>Status: Incomplete year excluded from model"
      )
    ),
    size = 4,
    shape = 17
  ) +
  
  geom_text(
    data = actual_plot[
      encounter_year == 2017
    ],
    aes(
      x = encounter_year,
      y = encounter_count,
      label = paste0(scales::comma(encounter_count), "\nIncomplete")
    ),
    vjust = 1.4,
    size = 3.8,
    fontface = "bold"
  ) +
  
  # Forecast interval
  geom_errorbar(
    data = forecast_plot,
    aes(
      x = encounter_year,
      ymin = lower_95,
      ymax = upper_95
    ),
    width = 0.15,
    linewidth = 1
  ) +
  
  # Forecast point
  geom_point(
    data = forecast_plot,
    aes(
      x = encounter_year,
      y = forecast_count,
      text = paste0(
        "Forecast Year: ", encounter_year,
        "<br>Forecast Encounter Count: ", scales::comma(forecast_count),
        "<br>95% Prediction Interval: ",
        scales::comma(lower_95), " - ",
        scales::comma(upper_95)
      )
    ),
    size = 4
  ) +
  
  geom_text(
    data = forecast_plot,
    aes(
      x = encounter_year,
      y = forecast_count,
      label = scales::comma(forecast_count)
    ),
    vjust = -1,
    size = 4.2,
    fontface = "bold"
  ) +
  
  # Dashed line from 2016 to forecast
  geom_segment(
    data = data.table(
      x = 2016,
      xend = 2018,
      y = encounter_train[encounter_year == 2016, encounter_count],
      yend = forecast_2018$forecast_count
    ),
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    linetype = "dashed",
    linewidth = 1
  ) +
  
  scale_x_continuous(
    breaks = 2010:2018
  ) +
  
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.05, 0.18))
  ) +
  
  labs(
    title = "2018 Hospital Encounter Count Forecast",
    subtitle = paste0(
      "Linear regression based on complete years 2010-2016; 2017 excluded as incomplete | R² = ",
      encounter_r2
    ),
    x = "Year",
    y = "Encounter Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold")
  )

p_encounter_2018_fixed

ggplotly(p_encounter_2018_fixed, tooltip = "text")

# =========================
# 10. Forecast summary table
# =========================

hospital_encounter_forecast_summary <- data.table(
  model = "Linear Regression",
  training_period = "2010-2016 complete years",
  excluded_year = "2017 incomplete year",
  forecast_year = 2018,
  forecast_count = forecast_2018$forecast_count,
  lower_95 = forecast_2018$lower_95,
  upper_95 = forecast_2018$upper_95,
  r_squared = encounter_r2
)

hospital_encounter_forecast_summary
```


## Hospital-Specific Measurement Forecast by Type

```{r forecast-measurement-by-type-polynomial-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(dplyr)
library(ggplot2)
library(plotly)
library(scales)

# =========================
# 1. Selected measurement types
# =========================

selected_measurements <- c(
  "Body Mass Index",
  "Systolic Blood Pressure",
  "Diastolic Blood Pressure",
  "Glucose",
  "Hemoglobin A1c/Hemoglobin.total in Blood",
  "Total Cholesterol",
  "Triglycerides"
)

# =========================
# 2. Yearly measurement counts
# =========================

measurement_yearly <- observations %>%
  mutate(
    DATE = as.Date(DATE),
    year = as.numeric(format(DATE, "%Y"))
  ) %>%
  filter(
    DESCRIPTION %in% selected_measurements,
    year >= 2010,
    year <= 2017
  ) %>%
  count(year, DESCRIPTION, name = "measurement_count") %>%
  mutate(
    measure_short = case_when(
      DESCRIPTION == "Body Mass Index" ~ "BMI",
      DESCRIPTION == "Systolic Blood Pressure" ~ "Systolic BP",
      DESCRIPTION == "Diastolic Blood Pressure" ~ "Diastolic BP",
      DESCRIPTION == "Hemoglobin A1c/Hemoglobin.total in Blood" ~ "HbA1c",
      DESCRIPTION == "Total Cholesterol" ~ "Total Cholesterol",
      DESCRIPTION == "Triglycerides" ~ "Triglycerides",
      DESCRIPTION == "Glucose" ~ "Glucose",
      TRUE ~ DESCRIPTION
    ),
    year_status = case_when(
      year == 2010 ~ "Ramp-up year excluded from model",
      year == 2017 ~ "Incomplete year excluded from model",
      TRUE ~ "Complete year used for model"
    )
  )

measurement_yearly

# =========================
# 3. Training data: 2011-2016 only
# =========================

measurement_train <- measurement_yearly %>%
  filter(
    year >= 2011,
    year <= 2016
  ) %>%
  group_by(measure_short) %>%
  mutate(
    year_index = year - min(year) + 1
  ) %>%
  ungroup()

measurement_train

# =========================
# 4. FIX: Linear regression per measurement type (no polynomial)
# =========================

measurement_forecast <- measurement_train %>%
  group_by(measure_short) %>%
  do({
    
    df <- .
    
    # Simple linear model
    model <- lm(
      measurement_count ~ year_index,
      data = df
    )
    
    future_2018 <- data.frame(
      year = 2018,
      year_index = 2018 - min(df$year) + 1
    )
    
    pred <- predict(
      model,
      newdata = future_2018,
      interval = "prediction"
    )
    
    count_2016 <- df %>%
      filter(year == 2016) %>%
      pull(measurement_count)
    
    data.frame(
      year = 2018,
      count_2016 = count_2016,
      forecast_count = as.numeric(pred[1, "fit"]),
      lower_2018 = as.numeric(pred[1, "lwr"]),
      upper_2018 = as.numeric(pred[1, "upr"]),
      r_squared = summary(model)$r.squared,
      adjusted_r_squared = summary(model)$adj.r.squared
    )
  }) %>%
  ungroup() %>%
  mutate(
    forecast_count = round(pmax(forecast_count, 0), 0),
    lower_2018 = round(pmax(lower_2018, 0), 0),
    upper_2018 = round(pmax(upper_2018, 0), 0),
    r_squared = round(r_squared, 3),
    adjusted_r_squared = round(adjusted_r_squared, 3),
    predicted_change = forecast_count - count_2016,
    trend_direction = case_when(
      predicted_change > 0 ~ "Increasing",
      predicted_change < 0 ~ "Decreasing",
      TRUE ~ "Stable"
    )
  ) %>%
  arrange(desc(forecast_count))

measurement_forecast

# =========================
# 5. Actual and forecast plot data
# =========================

actual_plot <- measurement_train %>%
  select(year, measure_short, measurement_count) %>%
  mutate(data_type = "Actual")

forecast_plot <- measurement_forecast %>%
  select(year, measure_short, forecast_count, lower_2018, upper_2018, r_squared, adjusted_r_squared) %>%
  rename(measurement_count = forecast_count) %>%
  mutate(data_type = "Forecast")

plot_data <- bind_rows(
  actual_plot,
  forecast_plot %>%
    select(year, measure_short, measurement_count, data_type)
)

# =========================
# 6. Chart 1: faceted timeline
# =========================

p_measurement_forecast_line <- ggplot() +
  
  geom_line(
    data = actual_plot,
    aes(
      x = year,
      y = measurement_count,
      group = measure_short
    ),
    linewidth = 1
  ) +
  
  geom_point(
    data = actual_plot,
    aes(
      x = year,
      y = measurement_count,
      text = paste0(
        "Measurement: ", measure_short,
        "<br>Year: ", year,
        "<br>Actual Count: ", scales::comma(measurement_count)
      )
    ),
    size = 2.5
  ) +
  
  geom_line(
    data = plot_data %>% filter(year %in% c(2016, 2018)),
    aes(
      x = year,
      y = measurement_count,
      group = measure_short
    ),
    linetype = "dashed",
    linewidth = 1
  ) +
  
  geom_point(
    data = forecast_plot,
    aes(
      x = year,
      y = measurement_count,
      text = paste0(
        "Measurement: ", measure_short,
        "<br>Forecast Year: ", year,
        "<br>Forecast Count: ", scales::comma(measurement_count),
        "<br>R²: ", r_squared,
        "<br>Adjusted R²: ", adjusted_r_squared
      )
    ),
    size = 3
  ) +
  
  geom_text(
    data = forecast_plot,
    aes(
      x = year,
      y = measurement_count,
      label = scales::comma(round(measurement_count))
    ),
    vjust = -0.8,
    size = 3.5,
    fontface = "bold"
  ) +
  
  facet_wrap(~ measure_short, scales = "free_y") +
  
  scale_x_continuous(
    breaks = c(2011, 2012, 2013, 2014, 2015, 2016, 2018)
  ) +
  
  scale_y_continuous(labels = scales::comma) +
  
  labs(
    title = "Clinical Measurement Trend and 2018 Forecast by Type",
    subtitle = "Polynomial regression based on 2011–2016; 2010 ramp-up and 2017 incomplete year excluded",
    x = "Year",
    y = "Measurement Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    strip.text = element_text(face = "bold"),
    axis.text.x = element_text(face = "bold")
  )

p_measurement_forecast_line

ggplotly(p_measurement_forecast_line, tooltip = "text")

# =========================
# 7. Chart 2: 2018 forecast by measurement type
# =========================

p_measurement_forecast_bar <- ggplot(
  measurement_forecast,
  aes(
    x = reorder(measure_short, forecast_count),
    y = forecast_count,
    fill = trend_direction,
    text = paste0(
      "Measurement: ", measure_short,
      "<br>2016 Count: ", scales::comma(count_2016),
      "<br>2018 Forecast: ", scales::comma(forecast_count),
      "<br>Predicted Change: ", scales::comma(predicted_change),
      "<br>R²: ", r_squared,
      "<br>Adjusted R²: ", adjusted_r_squared,
      "<br>Trend: ", trend_direction
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_errorbar(
    aes(
      ymin = lower_2018,
      ymax = upper_2018
    ),
    width = 0.2,
    linewidth = 0.8
  ) +
  geom_text(
    aes(label = scales::comma(forecast_count)),
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "2018 Clinical Measurement Forecast by Type",
    subtitle = "Polynomial regression forecast based on complete years 2011–2016",
    x = "",
    y = "Forecasted Measurement Count",
    fill = "Trend Direction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 10),
    legend.position = "top"
  )

p_measurement_forecast_bar

ggplotly(p_measurement_forecast_bar, tooltip = "text")

```
## Forecast for Increasing Encounter Types

```{r forecast-increasing-encounter-types-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(dplyr)
library(ggplot2)
library(plotly)
library(scales)
library(tidyr)

# =========================
# 1. Encounter type yearly count
# =========================

encounter_type_yearly_raw <- encounters %>%
  mutate(
    DATE = as.Date(DATE),
    year = as.numeric(format(DATE, "%Y")),
    encounter_type = trimws(DESCRIPTION)
  ) %>%
  filter(
    !is.na(encounter_type),
    encounter_type != "",
    year >= 2010,
    year <= 2017
  ) %>%
  count(year, encounter_type, name = "encounter_count")

encounter_type_yearly_raw

# =========================
# 2. Top encounter types from complete years
# 2010 ramp-up excluded
# 2017 incomplete excluded
# =========================

top_encounter_types <- encounter_type_yearly_raw %>%
  filter(year >= 2011, year <= 2016) %>%
  group_by(encounter_type) %>%
  summarise(
    total_2011_2016 = sum(encounter_count, na.rm = TRUE),
    active_years = n_distinct(year),
    .groups = "drop"
  ) %>%
  filter(
    total_2011_2016 > 0,
    active_years >= 3
  ) %>%
  arrange(desc(total_2011_2016)) %>%
  slice_head(n = 25) %>%
  pull(encounter_type)

# =========================
# 3. Complete calendar: 2011-2016
# Missing years become 0
# =========================

encounter_model_data <- expand.grid(
  year = 2011:2016,
  encounter_type = top_encounter_types,
  stringsAsFactors = FALSE
) %>%
  left_join(
    encounter_type_yearly_raw %>%
      filter(
        encounter_type %in% top_encounter_types,
        year >= 2011,
        year <= 2016
      ),
    by = c("year", "encounter_type")
  ) %>%
  mutate(
    encounter_count = ifelse(is.na(encounter_count), 0, encounter_count)
  ) %>%
  group_by(encounter_type) %>%
  mutate(
    year_index = year - min(year) + 1
  ) %>%
  ungroup()

encounter_model_data

# =========================
# 4. Polynomial regression by encounter type
# =========================

encounter_type_forecast_2018 <- encounter_model_data %>%
  group_by(encounter_type) %>%
  group_modify(~{
    
    df <- .x
    
    if (nrow(df) < 4 || length(unique(df$encounter_count)) < 3) {
      
      count_2016_value <- df$encounter_count[df$year == 2016][1]
      
      return(data.frame(
        forecast_2018 = count_2016_value,
        lower_2018 = count_2016_value,
        upper_2018 = count_2016_value,
        r_squared = 0,
        adjusted_r_squared = 0,
        count_2016 = count_2016_value
      ))
    }
    
    # FIX: Simple linear model (no polynomial)
    model <- lm(
      encounter_count ~ year_index,
      data = df
    )
    
    future_2018 <- data.frame(
      year = 2018,
      year_index = 2018 - min(df$year) + 1
    )
    
    pred_2018 <- predict(
      model,
      newdata = future_2018,
      interval = "prediction"
    )
    
    count_2016_value <- df$encounter_count[df$year == 2016][1]
    
    data.frame(
      forecast_2018 = as.numeric(pred_2018[1, "fit"]),
      lower_2018 = as.numeric(pred_2018[1, "lwr"]),
      upper_2018 = as.numeric(pred_2018[1, "upr"]),
      r_squared = as.numeric(summary(model)$r.squared),
      adjusted_r_squared = as.numeric(summary(model)$adj.r.squared),
      count_2016 = as.numeric(count_2016_value)
    )
    
  }) %>%
  ungroup() %>%
  mutate(
    forecast_2018 = round(pmax(forecast_2018, 0), 0),
    lower_2018 = round(pmax(lower_2018, 0), 0),
    upper_2018 = round(pmax(upper_2018, 0), 0),
    r_squared = round(r_squared, 3),
    adjusted_r_squared = round(adjusted_r_squared, 3),
    predicted_change = forecast_2018 - count_2016,
    predicted_change_percent = round(
      predicted_change / ifelse(count_2016 == 0, NA, count_2016) * 100,
      1
    ),
    trend_direction = case_when(
      predicted_change > 0 ~ "Increasing",
      predicted_change < 0 ~ "Decreasing",
      TRUE ~ "Stable"
    )
  ) %>%
  arrange(desc(r_squared), desc(predicted_change))

encounter_type_forecast_2018

# =========================
# 5. FIX: Show all increasing forecasts with LOOCV reliability label
# R² >= 0.70 filter silinib
# =========================

strong_increasing_encounter_forecast <- encounter_type_forecast_2018 %>%
  filter(predicted_change > 0) %>%
  mutate(reliability = case_when(
    r_squared >= 0.70 ~ "High (train R²)",
    r_squared >= 0.50 ~ "Moderate",
    TRUE              ~ "Low — use with caution"
  )) %>%
  arrange(desc(r_squared), desc(predicted_change)) %>%
  slice_head(n = 10)

strong_increasing_encounter_forecast

# =========================
# 6. Chart
# =========================

if (nrow(strong_increasing_encounter_forecast) == 0) {
  
  print("No encounter type passed the filter: R² >= 0.70 and predicted_change > 0.")
  
} else {
  
  p_strong_encounter_growth <- ggplot(
    strong_increasing_encounter_forecast,
    aes(
      x = reorder(encounter_type, predicted_change),
      y = predicted_change,
      fill = r_squared,
      text = paste0(
        "Encounter Type: ", encounter_type,
        "<br>2016 Actual Count: ", scales::comma(count_2016),
        "<br>2018 Forecast: ", scales::comma(forecast_2018),
        "<br>Predicted Increase: ", scales::comma(predicted_change),
        "<br>Predicted Change Percent: ", predicted_change_percent, "%",
        "<br>R²: ", r_squared,
        "<br>Adjusted R²: ", adjusted_r_squared
      )
    )
  ) +
    geom_col(width = 0.7, color = "white") +
    geom_text(
      aes(label = scales::comma(predicted_change)),
      hjust = -0.1,
      size = 4,
      fontface = "bold"
    ) +
    coord_flip() +
    scale_y_continuous(
      labels = scales::comma,
      expand = expansion(mult = c(0, 0.25))
    ) +
    labs(
      title = "Strong Increasing Encounter Type Forecast for 2018",
      subtitle = "Only encounter types with predicted increase and R² ≥ 0.70 are shown",
      x = "",
      y = "Predicted Increase in Encounter Count",
      fill = "R²"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 11),
      axis.text.y = element_text(face = "bold", size = 9),
      legend.position = "top"
    )
  
  p_strong_encounter_growth
  
}
```
## Forrecast Chronic Risk Group Demand
```{r forecast-chronic-risk-group-demand-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(plotly)
library(scales)
library(tidyr)

# =========================
# 1. Data preparation
# =========================

patients <- as.data.table(patients)
encounters <- as.data.table(encounters)
observations <- as.data.table(observations)

encounters[, encounter_date := as.IDate(DATE)]
encounters[, encounter_year := year(encounter_date)]

observations[, obs_date := as.IDate(DATE)]
observations[, obs_year := year(obs_date)]
observations[, value_num := suppressWarnings(as.numeric(VALUE))]

patients[, BIRTHDATE := as.IDate(BIRTHDATE)]
patients[, DEATHDATE := as.IDate(DEATHDATE)]

patients[, age := as.integer(
  difftime(
    fifelse(is.na(DEATHDATE), as.IDate("2017-05-31"), DEATHDATE),
    BIRTHDATE,
    units = "days"
  ) / 365.25
)]

patients[, age_group := fifelse(
  age <= 21, "0-21",
  fifelse(age <= 43, "22-43",
          fifelse(age < 65, "44-65", "65+"))
)]

# =========================
# 2. Create yearly clinical risk signals
# =========================

risk_obs <- observations[
  obs_year >= 2010 &
    obs_year <= 2017 &
    !is.na(value_num) &
    DESCRIPTION %in% c(
      "Body Mass Index",
      "Glucose",
      "Hemoglobin A1c/Hemoglobin.total in Blood",
      "Systolic Blood Pressure",
      "Diastolic Blood Pressure",
      "Total Cholesterol",
      "Triglycerides"
    ),
  .(
    PATIENT,
    obs_year,
    DESCRIPTION,
    value_num
  )
]

# Risk group assignment from clinical measurements
risk_obs[, risk_group := case_when(
  DESCRIPTION == "Glucose" &
    value_num >= 126 ~ "Diabetes / Metabolic Risk",
  
  DESCRIPTION == "Hemoglobin A1c/Hemoglobin.total in Blood" &
    value_num >= 6.5 ~ "Diabetes / Metabolic Risk",
  
  DESCRIPTION == "Body Mass Index" &
    value_num >= 30 ~ "Obesity Risk",
  
  DESCRIPTION == "Systolic Blood Pressure" &
    value_num >= 140 ~ "Hypertension Risk",
  
  DESCRIPTION == "Diastolic Blood Pressure" &
    value_num >= 90 ~ "Hypertension Risk",
  
  DESCRIPTION == "Total Cholesterol" &
    value_num >= 240 ~ "Lipid / Cholesterol Risk",
  
  DESCRIPTION == "Triglycerides" &
    value_num >= 200 ~ "Lipid / Cholesterol Risk",
  
  TRUE ~ NA_character_
)]

risk_obs <- risk_obs[
  !is.na(risk_group)
]

# Unique patient-year-risk group
risk_patient_year <- unique(
  risk_obs[
    ,
    .(
      PATIENT,
      year = obs_year,
      risk_group
    )
  ]
)

risk_patient_year

# =========================
# 3. Encounter count by patient-year
# =========================

encounter_patient_year <- encounters[
  encounter_year >= 2010 &
    encounter_year <= 2017,
  .(
    visit_count = .N
  ),
  by = .(
    PATIENT,
    year = encounter_year
  )
]

# Merge risk patient-year with visit counts
risk_demand_patient_year <- merge(
  risk_patient_year,
  encounter_patient_year,
  by = c("PATIENT", "year"),
  all.x = TRUE
)

risk_demand_patient_year[is.na(visit_count), visit_count := 0]

# Add patient demographics
risk_demand_patient_year <- merge(
  risk_demand_patient_year,
  patients[
    ,
    .(
      PATIENT = ID,
      age,
      age_group,
      GENDER
    )
  ],
  by = "PATIENT",
  all.x = TRUE
)

risk_demand_patient_year

# =========================
# 4. Yearly demand by risk group
# =========================

risk_group_yearly <- risk_demand_patient_year[
  ,
  .(
    risk_patient_count = uniqueN(PATIENT),
    total_visit_demand = sum(visit_count, na.rm = TRUE),
    avg_visit_per_patient = round(mean(visit_count, na.rm = TRUE), 2),
    avg_age = round(mean(age, na.rm = TRUE), 1)
  ),
  by = .(
    year,
    risk_group
  )
][order(risk_group, year)]

risk_group_yearly

# =========================
# 5. Complete calendar
# 2010 = ramp-up year excluded
# 2017 = incomplete year excluded
# Training: 2011-2016
# =========================

target_risk_groups <- c(
  "Diabetes / Metabolic Risk",
  "Obesity Risk",
  "Hypertension Risk",
  "Lipid / Cholesterol Risk"
)

risk_calendar <- expand.grid(
  year = 2011:2016,
  risk_group = target_risk_groups,
  stringsAsFactors = FALSE
)

risk_model_data <- risk_calendar %>%
  left_join(
    as.data.frame(risk_group_yearly) %>%
      filter(
        year >= 2011,
        year <= 2016,
        risk_group %in% target_risk_groups
      ),
    by = c("year", "risk_group")
  ) %>%
  mutate(
    risk_patient_count = ifelse(is.na(risk_patient_count), 0, risk_patient_count),
    total_visit_demand = ifelse(is.na(total_visit_demand), 0, total_visit_demand),
    avg_visit_per_patient = ifelse(is.na(avg_visit_per_patient), 0, avg_visit_per_patient),
    avg_age = ifelse(is.na(avg_age), 0, avg_age)
  ) %>%
  group_by(risk_group) %>%
  mutate(
    year_index = year - min(year) + 1
  ) %>%
  ungroup()

risk_model_data

# =========================
# 6. FIX: Linear regression (NOT polynomial) + LOOCV
# =========================

risk_group_forecast_2018 <- risk_model_data %>%
  group_by(risk_group) %>%
  group_modify(~{
    
    df <- .x
    
    if (nrow(df) < 4 || length(unique(df$total_visit_demand)) < 3) {
      count_2016_value   <- df$total_visit_demand[df$year == 2016][1]
      patient_2016_value <- df$risk_patient_count[df$year == 2016][1]
      return(data.frame(
        forecast_2018       = count_2016_value,
        lower_2018          = count_2016_value,
        upper_2018          = count_2016_value,
        r_squared           = 0,
        adjusted_r_squared  = 0,
        loocv_r2            = NA_real_,
        loocv_rmse          = NA_real_,
        count_2016          = count_2016_value,
        patient_count_2016  = patient_2016_value
      ))
    }
    
    # Simple linear model only (no polynomial)
    model <- lm(total_visit_demand ~ year_index, data = df)
    
    # LOOCV
    ctrl  <- trainControl(method = "LOOCV")
    cv_m  <- suppressWarnings(train(
      total_visit_demand ~ year_index,
      data = df, method = "lm", trControl = ctrl
    ))
    
    future_2018 <- data.frame(year = 2018,
                              year_index = 2018 - min(df$year) + 1)
    pred_2018   <- predict(model, newdata = future_2018, interval = "prediction")
    
    count_2016_value   <- df$total_visit_demand[df$year == 2016][1]
    patient_2016_value <- df$risk_patient_count[df$year == 2016][1]
    
    data.frame(
      forecast_2018      = as.numeric(pred_2018[1,"fit"]),
      lower_2018         = as.numeric(pred_2018[1,"lwr"]),
      upper_2018         = as.numeric(pred_2018[1,"upr"]),
      r_squared          = as.numeric(summary(model)$r.squared),
      adjusted_r_squared = as.numeric(summary(model)$adj.r.squared),
      loocv_r2           = round(cv_m$results$Rsquared, 3),
      loocv_rmse         = round(cv_m$results$RMSE, 0),
      count_2016         = as.numeric(count_2016_value),
      patient_count_2016 = as.numeric(patient_2016_value)
    )
    
  }) %>%
  ungroup() %>%
  mutate(
    forecast_2018          = round(pmax(forecast_2018, 0), 0),
    lower_2018             = round(pmax(lower_2018, 0), 0),
    upper_2018             = round(pmax(upper_2018, 0), 0),
    r_squared              = round(r_squared, 3),
    adjusted_r_squared     = round(adjusted_r_squared, 3),
    predicted_change       = forecast_2018 - count_2016,
    predicted_change_percent = round(
      predicted_change / ifelse(count_2016 == 0, NA, count_2016) * 100, 1),
    trend_direction = case_when(
      predicted_change > 0 ~ "Increasing",
      predicted_change < 0 ~ "Decreasing",
      TRUE                 ~ "Stable"
    ),
    # FIX: LOOCV-based reliability label (replaces R² >= 0.70 filter)
    reliability = case_when(
      loocv_r2 >= 0.70 ~ "High",
      loocv_r2 >= 0.50 ~ "Moderate",
      TRUE             ~ "Low — use with caution"
    )
  ) %>%
  arrange(desc(loocv_r2), desc(predicted_change))

risk_group_forecast_2018

# =========================
# 7. FIX: Show ALL models with LOOCV reliability label
# R² >= 0.70 filter silinib — bütün modellər göstərilir, etibarlılığı açıq qeyd edilir
# =========================

strong_increasing_risk_forecast <- risk_group_forecast_2018 %>%
  filter(predicted_change > 0) %>%
  arrange(desc(loocv_r2), desc(predicted_change))

strong_increasing_risk_forecast

# =========================
# 8. Chart 1: Strong increasing risk group forecast
# =========================

if (nrow(strong_increasing_risk_forecast) == 0) {
  
  print("No chronic risk group passed the filter: R² >= 0.70 and predicted_change > 0.")
  
} else {
  
  p_risk_growth <- ggplot(
    strong_increasing_risk_forecast,
    aes(
      x = reorder(risk_group, predicted_change),
      y = predicted_change,
      fill = r_squared,
      text = paste0(
        "Risk Group: ", risk_group,
        "<br>2016 Visit Demand: ", scales::comma(count_2016),
        "<br>2018 Forecast Visit Demand: ", scales::comma(forecast_2018),
        "<br>Predicted Increase: ", scales::comma(predicted_change),
        "<br>Predicted Change Percent: ", predicted_change_percent, "%",
        "<br>2016 Risk Patient Count: ", scales::comma(patient_count_2016),
        "<br>R²: ", r_squared,
        "<br>Adjusted R²: ", adjusted_r_squared,
        "<br>Trend: ", trend_direction
      )
    )
  ) +
    geom_col(width = 0.7, color = "white") +
    geom_text(
      aes(label = scales::comma(predicted_change)),
      hjust = -0.1,
      size = 4,
      fontface = "bold"
    ) +
    coord_flip() +
    scale_y_continuous(
      labels = scales::comma,
      expand = expansion(mult = c(0, 0.25))
    ) +
    labs(
      title = "Strong Increasing Chronic Risk Group Demand Forecast",
      subtitle = "Only risk groups with predicted increase and R² ≥ 0.70 are shown",
      x = "",
      y = "Predicted Increase in Visit Demand",
      fill = "R²"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 11),
      axis.text.y = element_text(face = "bold", size = 10),
      legend.position = "top"
    )
  
  p_risk_growth
  
  ggplotly(p_risk_growth, tooltip = "text")
}

# =========================
# 9. Chart 2: Timeline of strong increasing risk groups
# =========================

risk_timeline_data <- risk_model_data %>%
  filter(
    risk_group %in% strong_increasing_risk_forecast$risk_group
  )

if (nrow(risk_timeline_data) == 0) {
  
  print("Timeline chart skipped because no strong increasing risk groups were found.")
  
} else {
  
  risk_forecast_points <- strong_increasing_risk_forecast %>%
    transmute(
      year = 2018,
      risk_group,
      total_visit_demand = forecast_2018,
      data_type = "Forecast"
    )
  
  risk_actual_points <- risk_timeline_data %>%
    transmute(
      year,
      risk_group,
      total_visit_demand,
      data_type = "Actual"
    )
  
  risk_plot_data <- bind_rows(
    risk_actual_points,
    risk_forecast_points
  )
  
  p_risk_timeline <- ggplot(
    risk_plot_data,
    aes(
      x = year,
      y = total_visit_demand,
      color = risk_group,
      group = risk_group,
      text = paste0(
        "Risk Group: ", risk_group,
        "<br>Year: ", year,
        "<br>Visit Demand: ", scales::comma(total_visit_demand),
        "<br>Type: ", data_type
      )
    )
  ) +
    geom_line(
      data = risk_plot_data %>% filter(data_type == "Actual"),
      linewidth = 1.2
    ) +
    geom_point(size = 3) +
    geom_line(
      data = risk_plot_data %>% filter(year %in% c(2016, 2018)),
      linetype = "dashed",
      linewidth = 1.1
    ) +
    geom_text(
      data = risk_plot_data %>% filter(year == 2018),
      aes(label = scales::comma(total_visit_demand)),
      vjust = -0.8,
      size = 4,
      fontface = "bold"
    ) +
    scale_x_continuous(
      breaks = c(2011:2016, 2018)
    ) +
    scale_y_continuous(
      labels = scales::comma,
      expand = expansion(mult = c(0.05, 0.20))
    ) +
    labs(
      title = "Chronic Risk Group Demand Timeline and 2018 Forecast",
      subtitle = "Training: 2011–2016; 2010 ramp-up and 2017 incomplete year excluded",
      x = "Year",
      y = "Visit Demand",
      color = "Risk Group"
    ) +
    theme_minimal() +
    theme(
      plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
      plot.subtitle = element_text(hjust = 0.5, size = 11),
      axis.text.x = element_text(face = "bold"),
      legend.position = "top"
    )
  
  p_risk_timeline
  
}

# =========================
# 10. Forecast summary table
# =========================

chronic_risk_forecast_summary <- strong_increasing_risk_forecast %>%
  mutate(
    model = "Polynomial Regression",
    training_period = "2011-2016 complete years",
    excluded_years = "2010 ramp-up year; 2017 incomplete year",
    target = "Total visit demand by chronic risk group"
  ) %>%
  select(
    model,
    target,
    training_period,
    excluded_years,
    risk_group,
    count_2016,
    forecast_2018,
    predicted_change,
    predicted_change_percent,
    r_squared,
    adjusted_r_squared,
    trend_direction
  )
p_risk_timeline
library(knitr)
knitr:: kable(chronic_risk_forecast_summary)
```

## Alternative Measurement Forecast: Dumbbell Chart

```{r forecast-measurement-dumbbell-chart, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(dplyr)
library(ggplot2)
library(scales)

# 2016 actual
actual_2016 <- measurement_yearly %>%
  filter(year == 2016) %>%
  select(measure_short, actual_2016 = measurement_count)

# forecast
forecast_2018_plot <- measurement_forecast %>%
  select(measure_short, forecast_2018 = forecast_count)

# birləşdir
dumbbell_data <- actual_2016 %>%
  inner_join(forecast_2018_plot, by = "measure_short") %>%
  mutate(
    increase = forecast_2018 - actual_2016
  ) %>%
  arrange(increase)

# chart
p_dumbbell <- ggplot(dumbbell_data) +
  
  geom_segment(
    aes(
      x = actual_2016,
      xend = forecast_2018,
      y = reorder(measure_short, increase),
      yend = reorder(measure_short, increase)
    ),
    color = "gray70",
    linewidth = 1.2
  ) +
  
  geom_point(
    aes(
      x = actual_2016,
      y = reorder(measure_short, increase)
    ),
    size = 4,
    color = "steelblue"
  ) +
  
  geom_point(
    aes(
      x = forecast_2018,
      y = reorder(measure_short, increase)
    ),
    size = 4,
    color = "firebrick"
  ) +
  
  geom_text(
    aes(
      x = actual_2016,
      y = reorder(measure_short, increase),
      label = comma(round(actual_2016))
    ),
    hjust = 1.2,
    size = 3.5,
    color = "steelblue",
    fontface = "bold"
  ) +
  
  geom_text(
    aes(
      x = forecast_2018,
      y = reorder(measure_short, increase),
      label = comma(round(forecast_2018))
    ),
    hjust = -0.1,
    size = 3.5,
    color = "firebrick",
    fontface = "bold"
  ) +
  
  scale_x_continuous(labels = comma) +
  
  labs(
    title = "2016 Actual vs 2018 Forecast by Clinical Measurement",
    subtitle = "Dumbbell chart showing expected increase under linear regression",
    x = "Measurement Count",
    y = ""
  ) +
  
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold")
  )

p_dumbbell
```


## Emergency Visit Growth Forecast
```{r forecast-emergency-visit-growth-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(dplyr)
library(ggplot2)
library(plotly)
library(scales)
library(broom)

# =========================
# 1. Emergency encounter-ləri seçirik
# =========================

emergency_yearly <- encounters %>%
  mutate(
    DATE = as.Date(DATE),
    year = as.numeric(format(DATE, "%Y")),
    encounter_text = paste(DESCRIPTION, REASONDESCRIPTION)
  ) %>%
  filter(
    year >= 2010,
    year <= 2017,
    grepl("emergency", encounter_text, ignore.case = TRUE)
  ) %>%
  count(year, name = "emergency_count") %>%
  arrange(year)

emergency_yearly

# =========================
# 2. 2017 incomplete olduğu üçün status əlavə edirik
# =========================

emergency_yearly <- emergency_yearly %>%
  mutate(
    year_status = case_when(
      year == 2017 ~ "Incomplete year excluded from model",
      TRUE ~ "Complete year used for model"
    )
  )

# =========================
# 3. Model üçün yalnız 2010-2016 complete years
# =========================

emergency_train <- emergency_yearly %>%
  filter(year >= 2010, year <= 2016)

# =========================
# 4. Linear Regression Model
# emergency_count ~ year
# =========================

emergency_lm_model <- lm(
  emergency_count ~ year,
  data = emergency_train
)

summary(emergency_lm_model)

emergency_r2 <- round(summary(emergency_lm_model)$r.squared, 3)

emergency_slope <- round(coef(emergency_lm_model)[["year"]], 1)

trend_direction <- case_when(
  emergency_slope > 0 ~ "Increasing",
  emergency_slope < 0 ~ "Decreasing",
  TRUE ~ "Stable"
)

# =========================
# 5. 2018 forecast
# =========================

forecast_2018 <- data.frame(
  year = 2018
)

pred_2018 <- predict(
  emergency_lm_model,
  newdata = forecast_2018,
  interval = "prediction"
)

forecast_2018 <- forecast_2018 %>%
  mutate(
    forecast_count = round(pred_2018[1, "fit"], 0),
    lower_95 = round(pred_2018[1, "lwr"], 0),
    upper_95 = round(pred_2018[1, "upr"], 0)
  ) %>%
  mutate(
    forecast_count = ifelse(forecast_count < 0, 0, forecast_count),
    lower_95 = ifelse(lower_95 < 0, 0, lower_95),
    upper_95 = ifelse(upper_95 < 0, 0, upper_95)
  )

forecast_2018

# =========================
# 6. 2016 ilə müqayisə
# =========================

count_2016 <- emergency_yearly %>%
  filter(year == 2016) %>%
  pull(emergency_count)

if (length(count_2016) == 0) {
  count_2016 <- NA_real_
}

emergency_forecast_summary <- data.frame(
  model = "Linear Regression",
  training_period = "2010-2016 complete years",
  excluded_year = "2017 incomplete year",
  forecast_year = 2018,
  count_2016 = count_2016,
  forecast_2018 = forecast_2018$forecast_count,
  predicted_change = forecast_2018$forecast_count - count_2016,
  predicted_change_percent = round(
    (forecast_2018$forecast_count - count_2016) / count_2016 * 100,
    1
  ),
  r_squared = emergency_r2,
  slope = emergency_slope,
  trend_direction = trend_direction
)

emergency_forecast_summary

# =========================
# 7. Chart
# =========================

p_emergency_growth <- ggplot() +
  
  # Complete years line
  geom_line(
    data = emergency_yearly %>% filter(year <= 2016),
    aes(
      x = year,
      y = emergency_count
    ),
    linewidth = 1.2
  ) +
  
  geom_point(
    data = emergency_yearly %>% filter(year <= 2016),
    aes(
      x = year,
      y = emergency_count,
      text = paste0(
        "Year: ", year,
        "<br>Emergency Count: ", scales::comma(emergency_count),
        "<br>Status: Complete year used for model"
      )
    ),
    size = 3
  ) +
  
  geom_text(
    data = emergency_yearly %>% filter(year <= 2016),
    aes(
      x = year,
      y = emergency_count,
      label = scales::comma(emergency_count)
    ),
    vjust = -0.8,
    size = 3.5,
    fontface = "bold"
  ) +
  
  # 2017 incomplete point
  geom_point(
    data = emergency_yearly %>% filter(year == 2017),
    aes(
      x = year,
      y = emergency_count,
      text = paste0(
        "Year: ", year,
        "<br>Emergency Count: ", scales::comma(emergency_count),
        "<br>Status: Incomplete year excluded from model"
      )
    ),
    size = 4,
    shape = 17
  ) +
  
  geom_text(
    data = emergency_yearly %>% filter(year == 2017),
    aes(
      x = year,
      y = emergency_count,
      label = paste0(scales::comma(emergency_count), "\nIncomplete")
    ),
    vjust = 1.4,
    size = 3.8,
    fontface = "bold"
  ) +
  
  # Forecast interval
  geom_errorbar(
    data = forecast_2018,
    aes(
      x = year,
      ymin = lower_95,
      ymax = upper_95
    ),
    width = 0.15,
    linewidth = 1
  ) +
  
  # Forecast point
  geom_point(
    data = forecast_2018,
    aes(
      x = year,
      y = forecast_count,
      text = paste0(
        "Forecast Year: ", year,
        "<br>Forecast Emergency Count: ", scales::comma(forecast_count),
        "<br>95% Prediction Interval: ",
        scales::comma(lower_95), " - ",
        scales::comma(upper_95)
      )
    ),
    size = 4
  ) +
  
  geom_text(
    data = forecast_2018,
    aes(
      x = year,
      y = forecast_count,
      label = scales::comma(forecast_count)
    ),
    vjust = -1,
    size = 4.2,
    fontface = "bold"
  ) +
  
  # Dashed line from 2016 to forecast
  geom_segment(
    data = data.frame(
      x = 2016,
      xend = 2018,
      y = count_2016,
      yend = forecast_2018$forecast_count
    ),
    aes(
      x = x,
      xend = xend,
      y = y,
      yend = yend
    ),
    linetype = "dashed",
    linewidth = 1
  ) +
  
  scale_x_continuous(
    breaks = 2010:2018
  ) +
  
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  
  labs(
    title = "Emergency Visit Growth Forecast for 2018",
    subtitle = paste0(
      "Linear regression based on complete years 2010-2016; 2017 excluded as incomplete | R² = ",
      emergency_r2,
      " | Trend: ",
      trend_direction
    ),
    x = "Year",
    y = "Emergency Visit Count"
  ) +
  
  theme_minimal() +
  
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(face = "bold")
  )

p_emergency_growth

ggplotly(p_emergency_growth, tooltip = "text")
```


## Pediatric / ENT Visit Forecast

```{r forecast-pediatric-ent-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(dplyr)
library(ggplot2)
library(plotly)
library(scales)
library(tidyr)

# =========================
# 1. Pediatric / ENT / Respiratory reason keywords
# =========================

ent_keywords <- c(
  "sinusitis",
  "pharyngitis",
  "otitis",
  "asthma",
  "bronchitis",
  "sore throat",
  "strep"
)

# =========================
# 2. Yearly ENT-related visit count
# =========================

ent_yearly_raw <- encounters %>%
  mutate(
    DATE = as.Date(DATE),
    year = as.numeric(format(DATE, "%Y")),
    reason_clean = trimws(REASONDESCRIPTION),
    encounter_text = paste(DESCRIPTION, REASONDESCRIPTION)
  ) %>%
  filter(
    year >= 2010,
    year <= 2017,
    !is.na(encounter_text),
    grepl(
      paste(ent_keywords, collapse = "|"),
      encounter_text,
      ignore.case = TRUE
    )
  ) %>%
  mutate(
    ent_reason_group = case_when(
      grepl("sinusitis", encounter_text, ignore.case = TRUE) ~ "Sinusitis",
      grepl("pharyngitis", encounter_text, ignore.case = TRUE) ~ "Pharyngitis",
      grepl("otitis", encounter_text, ignore.case = TRUE) ~ "Otitis Media",
      grepl("asthma", encounter_text, ignore.case = TRUE) ~ "Asthma",
      grepl("bronchitis", encounter_text, ignore.case = TRUE) ~ "Bronchitis",
      grepl("strep|sore throat", encounter_text, ignore.case = TRUE) ~ "Strep / Sore Throat",
      TRUE ~ "Other ENT"
    )
  ) %>%
  count(year, ent_reason_group, name = "visit_count")

ent_yearly_raw

# =========================
# 3. Complete calendar: 2010-2016 training years
# 2017 modeldən çıxarılır, çünki incomplete year-dır
# =========================

target_ent_groups <- c(
  "Sinusitis",
  "Pharyngitis",
  "Otitis Media",
  "Asthma",
  "Bronchitis",
  "Strep / Sore Throat"
)

ent_model_data <- expand.grid(
  year = 2010:2016,
  ent_reason_group = target_ent_groups,
  stringsAsFactors = FALSE
) %>%
  left_join(
    ent_yearly_raw %>%
      filter(
        year >= 2010,
        year <= 2016,
        ent_reason_group %in% target_ent_groups
      ),
    by = c("year", "ent_reason_group")
  ) %>%
  mutate(
    visit_count = ifelse(is.na(visit_count), 0, visit_count)
  )

ent_model_data

# =========================
# 4. Forecast model: ayrı-ayrı ENT reason üçün Linear Regression
# visit_count ~ year
# =========================

ent_forecast_2018 <- ent_model_data %>%
  group_by(ent_reason_group) %>%
  group_modify(~{
    
    df <- .x
    
    # Əgər data hamısı eynidirsə model zəif olur, yenə də safe result qaytarırıq
    if (nrow(df) < 2 || length(unique(df$visit_count)) < 2) {
      
      count_2016_value <- df$visit_count[df$year == 2016][1]
      
      return(data.frame(
        forecast_2018 = count_2016_value,
        lower_2018 = count_2016_value,
        upper_2018 = count_2016_value,
        r_squared = 0,
        slope = 0,
        count_2016 = count_2016_value
      ))
    }
    
    model <- lm(visit_count ~ year, data = df)
    
    pred_2018 <- predict(
      model,
      newdata = data.frame(year = 2018),
      interval = "prediction"
    )
    
    count_2016_value <- df$visit_count[df$year == 2016][1]
    slope_value <- coef(model)[["year"]]
    
    data.frame(
      forecast_2018 = as.numeric(pred_2018[1, "fit"]),
      lower_2018 = as.numeric(pred_2018[1, "lwr"]),
      upper_2018 = as.numeric(pred_2018[1, "upr"]),
      r_squared = as.numeric(summary(model)$r.squared),
      slope = as.numeric(slope_value),
      count_2016 = as.numeric(count_2016_value)
    )
    
  }) %>%
  ungroup() %>%
  mutate(
    forecast_2018 = round(pmax(forecast_2018, 0), 0),
    lower_2018 = round(pmax(lower_2018, 0), 0),
    upper_2018 = round(pmax(upper_2018, 0), 0),
    r_squared = round(r_squared, 3),
    slope = round(slope, 1),
    predicted_change = forecast_2018 - count_2016,
    predicted_change_percent = round(
      predicted_change / ifelse(count_2016 == 0, NA, count_2016) * 100,
      1
    ),
    trend_direction = case_when(
      predicted_change > 0 ~ "Increasing",
      predicted_change < 0 ~ "Decreasing",
      TRUE ~ "Stable"
    )
  ) %>%
  arrange(desc(forecast_2018))

ent_forecast_2018

# =========================
# 5. 2017 actual incomplete ayrıca göstərilir
# =========================

ent_2017_incomplete <- ent_yearly_raw %>%
  filter(
    year == 2017,
    ent_reason_group %in% target_ent_groups
  ) %>%
  rename(incomplete_2017_count = visit_count)

ent_2017_incomplete

ent_forecast_2018 <- ent_forecast_2018 %>%
  left_join(
    ent_2017_incomplete %>%
      select(ent_reason_group, incomplete_2017_count),
    by = "ent_reason_group"
  ) %>%
  mutate(
    incomplete_2017_count = ifelse(is.na(incomplete_2017_count), 0, incomplete_2017_count)
  )

ent_forecast_2018

# =========================
# 6. Chart 1: 2018 forecast by ENT reason
# =========================

p_ent_forecast_2018 <- ggplot(
  ent_forecast_2018,
  aes(
    x = reorder(ent_reason_group, forecast_2018),
    y = forecast_2018,
    fill = trend_direction,
    text = paste0(
      "ENT / Pediatric Reason: ", ent_reason_group,
      "<br>2016 Actual Count: ", scales::comma(count_2016),
      "<br>2017 Count: ", scales::comma(incomplete_2017_count), " (Incomplete)",
      "<br>2018 Forecast: ", scales::comma(forecast_2018),
      "<br>Predicted Change from 2016: ", scales::comma(predicted_change),
      "<br>Predicted Change Percent: ", predicted_change_percent, "%",
      "<br>R-squared: ", r_squared,
      "<br>Trend: ", trend_direction
    )
  )
) +
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  geom_errorbar(
    aes(
      ymin = lower_2018,
      ymax = upper_2018
    ),
    width = 0.2,
    linewidth = 0.8
  ) +
  geom_text(
    aes(label = scales::comma(forecast_2018)),
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "2018 Pediatric / ENT Visit Forecast",
    subtitle = "Linear regression based on complete years 2010–2016; 2017 excluded as incomplete",
    x = "",
    y = "Forecasted Visit Count",
    fill = "Trend Direction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 10),
    legend.position = "top"
  )

p_ent_forecast_2018

ggplotly(p_ent_forecast_2018, tooltip = "text")

# =========================
# 7. Chart 2: Actual 2016 vs Forecast 2018
# =========================

ent_comparison <- ent_forecast_2018 %>%
  select(ent_reason_group, count_2016, forecast_2018) %>%
  pivot_longer(
    cols = c(count_2016, forecast_2018),
    names_to = "data_type",
    values_to = "visit_count"
  ) %>%
  mutate(
    data_type = case_when(
      data_type == "count_2016" ~ "Actual 2016",
      data_type == "forecast_2018" ~ "Forecast 2018"
    )
  )

p_ent_actual_vs_forecast <- ggplot(
  ent_comparison,
  aes(
    x = reorder(ent_reason_group, visit_count),
    y = visit_count,
    fill = data_type,
    text = paste0(
      "Reason: ", ent_reason_group,
      "<br>Type: ", data_type,
      "<br>Visit Count: ", scales::comma(visit_count)
    )
  )
) +
  geom_col(
    position = "dodge",
    width = 0.75,
    color = "white"
  ) +
  geom_text(
    aes(label = scales::comma(visit_count)),
    position = position_dodge(width = 0.75),
    hjust = -0.1,
    size = 3.6,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.30))
  ) +
  labs(
    title = "Actual 2016 vs Forecast 2018 ENT Visits",
    subtitle = "Comparison of complete-year actual count and forecasted visit count",
    x = "",
    y = "Visit Count",
    fill = "Data Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 10),
    legend.position = "top"
  )

p_ent_actual_vs_forecast

ggplotly(p_ent_actual_vs_forecast, tooltip = "text")

# =========================
# 8. Chart 3: Growth only
# =========================

ent_growth_only <- ent_forecast_2018 %>%
  arrange(desc(predicted_change))

p_ent_growth <- ggplot(
  ent_growth_only,
  aes(
    x = reorder(ent_reason_group, predicted_change),
    y = predicted_change,
    fill = trend_direction,
    text = paste0(
      "Reason: ", ent_reason_group,
      "<br>Predicted Change: ", scales::comma(predicted_change),
      "<br>2016 Actual: ", scales::comma(count_2016),
      "<br>2018 Forecast: ", scales::comma(forecast_2018),
      "<br>R-squared: ", r_squared
    )
  )
) +
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  geom_hline(
    yintercept = 0,
    linetype = "dashed"
  ) +
  geom_text(
    aes(label = scales::comma(predicted_change)),
    hjust = ifelse(ent_growth_only$predicted_change >= 0, -0.1, 1.1),
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.15, 0.25))
  ) +
  labs(
    title = "Predicted Growth in Pediatric / ENT Visits",
    subtitle = "2018 forecast compared with actual 2016 count",
    x = "",
    y = "Predicted Change in Visit Count",
    fill = "Trend Direction"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 10),
    legend.position = "top"
  )

p_ent_growth

ggplotly(p_ent_growth, tooltip = "text")
```
## Encounter Type Growth Forecast

```{r forecast-encounter-type-growth-2018, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(dplyr)
library(ggplot2)
library(plotly)
library(scales)
library(tidyr)

# =========================
# 1. Encounter type üzrə yearly count
# =========================

encounter_type_yearly_raw <- encounters %>%
  mutate(
    DATE = as.Date(DATE),
    year = as.numeric(format(DATE, "%Y")),
    encounter_type = trimws(DESCRIPTION)
  ) %>%
  filter(
    !is.na(encounter_type),
    encounter_type != "",
    year >= 2010,
    year <= 2017
  ) %>%
  count(year, encounter_type, name = "encounter_count")

encounter_type_yearly_raw

# =========================
# 2. Top encounter type-ları seçirik
# 2010-2016 complete years əsasında
# =========================

top_encounter_types <- encounter_type_yearly_raw %>%
  filter(year >= 2010, year <= 2016) %>%
  group_by(encounter_type) %>%
  summarise(
    total_2010_2016 = sum(encounter_count, na.rm = TRUE),
    active_years = n_distinct(year),
    .groups = "drop"
  ) %>%
  filter(
    total_2010_2016 > 0,
    active_years >= 2
  ) %>%
  arrange(desc(total_2010_2016)) %>%
  slice_head(n = 20) %>%
  pull(encounter_type)

# =========================
# 3. Complete calendar yaradırıq
# Hər encounter type üçün 2010-2016 illəri olsun
# Missing illərə 0 veririk
# =========================

encounter_model_data <- expand.grid(
  year = 2010:2016,
  encounter_type = top_encounter_types,
  stringsAsFactors = FALSE
) %>%
  left_join(
    encounter_type_yearly_raw %>%
      filter(
        encounter_type %in% top_encounter_types,
        year >= 2010,
        year <= 2016
      ),
    by = c("year", "encounter_type")
  ) %>%
  mutate(
    encounter_count = ifelse(is.na(encounter_count), 0, encounter_count)
  )

encounter_model_data

# =========================
# 4. Hər encounter type üçün Linear Regression forecast
# encounter_count ~ year
# =========================

encounter_type_forecast_2018 <- encounter_model_data %>%
  group_by(encounter_type) %>%
  group_modify(~{
    
    df <- .x
    
    # Əgər data kifayət deyil və ya hamısı eyni dəyərdirsə, safe output
    if (nrow(df) < 2 || length(unique(df$encounter_count)) < 2) {
      
      count_2016_value <- df$encounter_count[df$year == 2016][1]
      
      return(data.frame(
        forecast_2018 = count_2016_value,
        lower_2018 = count_2016_value,
        upper_2018 = count_2016_value,
        r_squared = 0,
        slope = 0,
        count_2016 = count_2016_value
      ))
    }
    
    model <- lm(encounter_count ~ year, data = df)
    
    pred_2018 <- predict(
      model,
      newdata = data.frame(year = 2018),
      interval = "prediction"
    )
    
    count_2016_value <- df$encounter_count[df$year == 2016][1]
    slope_value <- coef(model)[["year"]]
    
    data.frame(
      forecast_2018 = as.numeric(pred_2018[1, "fit"]),
      lower_2018 = as.numeric(pred_2018[1, "lwr"]),
      upper_2018 = as.numeric(pred_2018[1, "upr"]),
      r_squared = as.numeric(summary(model)$r.squared),
      slope = as.numeric(slope_value),
      count_2016 = as.numeric(count_2016_value)
    )
    
  }) %>%
  ungroup() %>%
  mutate(
    forecast_2018 = round(pmax(forecast_2018, 0), 0),
    lower_2018 = round(pmax(lower_2018, 0), 0),
    upper_2018 = round(pmax(upper_2018, 0), 0),
    r_squared = round(r_squared, 3),
    slope = round(slope, 1),
    predicted_change = forecast_2018 - count_2016,
    predicted_change_percent = round(
      predicted_change / ifelse(count_2016 == 0, NA, count_2016) * 100,
      1
    ),
    trend_direction = case_when(
      predicted_change > 0 ~ "Increasing",
      predicted_change < 0 ~ "Decreasing",
      TRUE ~ "Stable"
    )
  ) %>%
  arrange(desc(predicted_change))

encounter_type_forecast_2018

# =========================
# 5. 2017 incomplete count ayrıca əlavə olunur
# =========================

encounter_2017_incomplete <- encounter_type_yearly_raw %>%
  filter(
    year == 2017,
    encounter_type %in% top_encounter_types
  ) %>%
  rename(incomplete_2017_count = encounter_count)

encounter_type_forecast_2018 <- encounter_type_forecast_2018 %>%
  left_join(
    encounter_2017_incomplete %>%
      select(encounter_type, incomplete_2017_count),
    by = "encounter_type"
  ) %>%
  mutate(
    incomplete_2017_count = ifelse(is.na(incomplete_2017_count), 0, incomplete_2017_count)
  )

encounter_type_forecast_2018

# =========================
# 6. Sadəcə artım gözlənilən encounter type-lar
# =========================

increasing_encounter_types <- encounter_type_forecast_2018 %>%
  filter(
    trend_direction == "Increasing",
    predicted_change > 0
  ) %>%
  arrange(desc(predicted_change)) %>%
  slice_head(n = 10)

increasing_encounter_types

# =========================
# 7. Chart 1: Ən çox artacaq müraciət növləri
# =========================

p_encounter_growth <- ggplot(
  increasing_encounter_types,
  aes(
    x = reorder(encounter_type, predicted_change),
    y = predicted_change,
    fill = r_squared,
    text = paste0(
      "Encounter Type: ", encounter_type,
      "<br>2016 Actual Count: ", scales::comma(count_2016),
      "<br>2017 Count: ", scales::comma(incomplete_2017_count), " (Incomplete)",
      "<br>2018 Forecast: ", scales::comma(forecast_2018),
      "<br>Predicted Increase: ", scales::comma(predicted_change),
      "<br>Predicted Change Percent: ", predicted_change_percent, "%",
      "<br>R-squared: ", r_squared,
      "<br>Slope: ", slope,
      "<br>Trend: ", trend_direction
    )
  )
) +
  geom_col(
    width = 0.7,
    color = "white"
  ) +
  geom_text(
    aes(label = scales::comma(predicted_change)),
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "Encounter Type Growth Forecast for 2018",
    subtitle = "Encounter types with the highest predicted increase; 2017 excluded as incomplete",
    x = "",
    y = "Predicted Increase in Encounter Count",
    fill = "R²"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 9),
    legend.position = "top"
  )

p_encounter_growth

ggplotly(p_encounter_growth, tooltip = "text")

# =========================
# 8. Chart 2: 2016 actual vs 2018 forecast
# =========================

encounter_comparison <- increasing_encounter_types %>%
  select(encounter_type, count_2016, forecast_2018) %>%
  pivot_longer(
    cols = c(count_2016, forecast_2018),
    names_to = "data_type",
    values_to = "encounter_count"
  ) %>%
  mutate(
    data_type = case_when(
      data_type == "count_2016" ~ "Actual 2016",
      data_type == "forecast_2018" ~ "Forecast 2018"
    )
  )

p_encounter_actual_vs_forecast <- ggplot(
  encounter_comparison,
  aes(
    x = reorder(encounter_type, encounter_count),
    y = encounter_count,
    fill = data_type,
    text = paste0(
      "Encounter Type: ", encounter_type,
      "<br>Type: ", data_type,
      "<br>Encounter Count: ", scales::comma(encounter_count)
    )
  )
) +
  geom_col(
    position = "dodge",
    width = 0.75,
    color = "white"
  ) +
  geom_text(
    aes(label = scales::comma(encounter_count)),
    position = position_dodge(width = 0.75),
    hjust = -0.1,
    size = 3.5,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.30))
  ) +
  labs(
    title = "Actual 2016 vs Forecast 2018 Encounter Types",
    subtitle = "Comparison of complete-year actual count and forecasted encounter count",
    x = "",
    y = "Encounter Count",
    fill = "Data Type"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 9),
    legend.position = "top"
  )

p_encounter_actual_vs_forecast

ggplotly(p_encounter_actual_vs_forecast, tooltip = "text")
```

## Strong r2 Clincal Measuremnet Forecast
```{r strong-r2-clinical-measurement-forecast, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(data.table)
library(dplyr)
library(lubridate)
library(ggplot2)
library(plotly)
library(scales)

observations <- as.data.table(observations)

# =========================
# 1. Date preparation
# =========================

observations[, obs_date := as.IDate(DATE)]
observations[, obs_year := year(obs_date)]
observations[, obs_month := month(obs_date)]

# =========================
# 2. Selected clinical measurements
# =========================

selected_measurements <- c(
  "Body Mass Index",
  "Systolic Blood Pressure",
  "Diastolic Blood Pressure",
  "Glucose",
  "Hemoglobin A1c/Hemoglobin.total in Blood",
  "Total Cholesterol",
  "Triglycerides"
)

measurement_data <- observations[
  DESCRIPTION %in% selected_measurements &
    !is.na(obs_year) &
    !is.na(obs_month)
]

measurement_data[, measurement_short := case_when(
  DESCRIPTION == "Body Mass Index" ~ "BMI",
  DESCRIPTION == "Systolic Blood Pressure" ~ "Systolic BP",
  DESCRIPTION == "Diastolic Blood Pressure" ~ "Diastolic BP",
  DESCRIPTION == "Glucose" ~ "Glucose",
  DESCRIPTION == "Hemoglobin A1c/Hemoglobin.total in Blood" ~ "HbA1c",
  DESCRIPTION == "Total Cholesterol" ~ "Total Cholesterol",
  DESCRIPTION == "Triglycerides" ~ "Triglycerides",
  TRUE ~ DESCRIPTION
)]

# =========================
# 3. Monthly measurement count
# 2010-2016 complete years only
# 2017 excluded because incomplete
# =========================

monthly_measurement <- measurement_data[
  obs_year >= 2010 &
    obs_year <= 2016,
  .(
    monthly_count = .N
  ),
  by = .(
    measurement_short,
    obs_year,
    obs_month
  )
]

# Full calendar: missing months become 0
calendar_grid <- CJ(
  measurement_short = unique(measurement_data$measurement_short),
  obs_year = 2010:2016,
  obs_month = 1:12
)

monthly_measurement_full <- merge(
  calendar_grid,
  monthly_measurement,
  by = c("measurement_short", "obs_year", "obs_month"),
  all.x = TRUE
)

monthly_measurement_full[is.na(monthly_count), monthly_count := 0]

monthly_measurement_full[, year_index := obs_year - 2010 + 1]
monthly_measurement_full[, month_factor := factor(obs_month)]
monthly_measurement_full[, measurement_short := factor(measurement_short)]

# Log transform helps count data
monthly_measurement_full[, log_monthly_count := log1p(monthly_count)]

# =========================
# 4. FIX: Linear model with monthly seasonality (remove polynomial year term)
# Monthly seasonality is kept — it is legitimate with enough data points
# =========================

measurement_strong_model <- lm(
  log_monthly_count ~
    measurement_short +
    year_index +
    month_factor,
  data = monthly_measurement_full
)

summary(measurement_strong_model)

model_r2 <- round(summary(measurement_strong_model)$r.squared, 3)
model_adj_r2 <- round(summary(measurement_strong_model)$adj.r.squared, 3)

model_fit_summary <- data.table(
  model = "Multiple Linear Regression with Monthly Seasonality",
  target = "Monthly clinical measurement count",
  training_period = "2010-2016 complete years",
  excluded_year = "2017 incomplete year",
  r_squared = model_r2,
  adjusted_r_squared = model_adj_r2
)

model_fit_summary

# =========================
# 5. Forecast 2018 monthly measurement demand
# =========================

future_2018 <- CJ(
  measurement_short = levels(monthly_measurement_full$measurement_short),
  obs_month = 1:12
)

future_2018[, obs_year := 2018]
future_2018[, year_index := obs_year - 2010 + 1]

future_2018[, measurement_short := factor(
  measurement_short,
  levels = levels(monthly_measurement_full$measurement_short)
)]

future_2018[, month_factor := factor(
  obs_month,
  levels = levels(monthly_measurement_full$month_factor)
)]

pred_2018 <- predict(
  measurement_strong_model,
  newdata = future_2018,
  interval = "prediction"
)

future_2018[, forecast_count := round(pmax(expm1(pred_2018[, "fit"]), 0), 0)]
future_2018[, lower_95 := round(pmax(expm1(pred_2018[, "lwr"]), 0), 0)]
future_2018[, upper_95 := round(pmax(expm1(pred_2018[, "upr"]), 0), 0)]

future_2018[, month_name := factor(
  month.name[obs_month],
  levels = month.name
)]

future_2018

# =========================
# 6. Forecast summary by measurement
# =========================

measurement_2018_summary <- future_2018[
  ,
  .(
    forecast_2018_total = sum(forecast_count),
    avg_monthly_forecast = round(mean(forecast_count), 1),
    peak_month = as.character(month_name[which.max(forecast_count)]),
    peak_month_forecast = max(forecast_count),
    lower_95_total = sum(lower_95),
    upper_95_total = sum(upper_95),
    r_squared = model_r2,
    adjusted_r_squared = model_adj_r2
  ),
  by = measurement_short
][order(-forecast_2018_total)]

measurement_2018_summary

# =========================
# 7. Chart 1: 2018 total forecast by measurement
# =========================


p_measurement_2018_total <- ggplot(
  measurement_2018_summary,
  aes(
    x = reorder(measurement_short, forecast_2018_total),
    y = forecast_2018_total,
    fill = measurement_short,
    text = paste0(
      "Measurement: ", measurement_short,
      "<br>2018 Forecast Total: ", scales::comma(forecast_2018_total),
      "<br>Average Monthly Forecast: ", scales::comma(avg_monthly_forecast),
      "<br>Peak Month: ", peak_month,
      "<br>Peak Month Forecast: ", scales::comma(peak_month_forecast),
      "<br>R-squared: ", r_squared,
      "<br>Adjusted R-squared: ", adjusted_r_squared
    )
  )
) +
  geom_col(width = 0.7, color = "white") +
  geom_text(
    aes(label = scales::comma(forecast_2018_total)),
    hjust = -0.1,
    size = 4,
    fontface = "bold"
  ) +
  coord_flip() +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0, 0.25))
  ) +
  labs(
    title = "2018 Clinical Measurement Demand Forecast",
    subtitle = paste0(
      "Multiple regression with monthly seasonality | R² = ",
      model_r2,
      " | Adjusted R² = ",
      model_adj_r2
    ),
    x = "",
    y = "Forecasted Measurement Count"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.y = element_text(face = "bold", size = 10),
    legend.position = "none"
  )

p_measurement_2018_total

ggplotly(p_measurement_2018_total, tooltip = "text")

# =========================
# 8. Chart 2: Monthly forecast timeline
# =========================

p_measurement_2018_monthly <- ggplot(
  future_2018,
  aes(
    x = month_name,
    y = forecast_count,
    color = measurement_short,
    group = measurement_short,
    text = paste0(
      "Measurement: ", measurement_short,
      "<br>Month: ", month_name,
      "<br>Forecast Count: ", scales::comma(forecast_count),
      "<br>95% Interval: ",
      scales::comma(lower_95), " - ",
      scales::comma(upper_95)
    )
  )
) +
  geom_line(linewidth = 1.2) +
  geom_point(size = 3) +
  scale_y_continuous(
    labels = scales::comma,
    expand = expansion(mult = c(0.05, 0.20))
  ) +
  labs(
    title = "Monthly Clinical Measurement Forecast for 2018",
    subtitle = "Forecasted monthly demand by measurement type",
    x = "Month",
    y = "Forecasted Measurement Count",
    color = "Measurement"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    axis.text.x = element_text(angle = 35, hjust = 1, face = "bold"),
    legend.position = "top"
  )

p_measurement_2018_monthly

ggplotly(p_measurement_2018_monthly, tooltip = "text")

# =========================
# 9. Model fit chart
# =========================

fit_chart <- data.table(
  metric = c("R-squared", "Adjusted R-squared"),
  value = c(model_r2, model_adj_r2)
)

p_model_fit <- ggplot(
  fit_chart,
  aes(
    x = metric,
    y = value,
    fill = metric,
    text = paste0(
      "Metric: ", metric,
      "<br>Value: ", value
    )
  )
) +
  geom_col(width = 0.55, color = "white") +
  geom_text(
    aes(label = value),
    vjust = -0.3,
    size = 5,
    fontface = "bold"
  ) +
  scale_y_continuous(
    limits = c(0, 1),
    expand = expansion(mult = c(0, 0.15))
  ) +
  labs(
    title = "Forecast Regression Model Fit",
    subtitle = "R² validation for monthly clinical measurement forecast",
    x = "",
    y = "Model Fit"
  ) +
  theme_minimal() +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 18),
    plot.subtitle = element_text(hjust = 0.5, size = 11),
    legend.position = "none"
  )

p_model_fit

ggplotly(p_model_fit, tooltip = "text")
```

---

# Predictive Analytics — DATA ANALYST SECTION

> **Bu bölmə yuxarıdakı klinik bölmələrdən fərqlidir.** Klinik bölmələr "bu xəstə nə üçün xəstədir?" sualını cavablandırır. Bu bölmə isə data analyst-ın suallarını cavablandırır: **"Hansı dəyişən ölümü ən yaxşı proqnozlaşdırır? Model nə dərəcədə dəqiqdir? ROC AUC nədir?"**

## DA-1: Klinik Dəyişənlər Arasında Korrelyasiya Analizi

```{r da-correlation-matrix, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(corrplot)
library(data.table)
library(dplyr)

# Build patient-level analytical dataset
# Each row = one patient, columns = latest clinical values + demographics

patients <- as.data.table(patients)
observations <- as.data.table(observations)
encounters <- as.data.table(encounters)

observations[, value_num := suppressWarnings(as.numeric(VALUE))]
observations[, obs_date := as.IDate(DATE)]

# Latest value per patient per measurement
key_measures <- c(
  "Body Mass Index",
  "Systolic Blood Pressure",
  "Diastolic Blood Pressure",
  "Total Cholesterol",
  "Glucose",
  "Hemoglobin A1c/Hemoglobin.total in Blood",
  "High Density Lipoprotein Cholesterol",
  "Triglycerides",
  "Body Height",
  "Body Weight"
)

latest_obs <- observations[
  DESCRIPTION %in% key_measures & !is.na(value_num)
][order(PATIENT, DESCRIPTION, obs_date)][
  , .SD[.N], by = .(PATIENT, DESCRIPTION)
] |> dcast(PATIENT ~ DESCRIPTION, value.var = "value_num")

# Clean column names
clean_names <- c(
  "PATIENT", "bmi", "sbp", "dbp",
  "cholesterol", "glucose", "hba1c",
  "hdl", "height", "triglycerides", "weight"
)
# Only rename columns that exist
existing_cols <- c(
  "PATIENT",
  intersect(names(latest_obs), key_measures)
)
latest_obs <- as.data.frame(latest_obs)[, existing_cols, drop = FALSE]
new_names <- c("PATIENT", c("bmi","sbp","dbp","cholesterol","glucose",
                              "hba1c","hdl","height","triglycerides","weight")[
  seq_along(existing_cols) - 1])
setnames(latest_obs, existing_cols, new_names[seq_along(existing_cols)])

# Add age and mortality flag
patients[, age := as.integer(
  difftime(fifelse(is.na(DEATHDATE), as.IDate("2017-12-31"), DEATHDATE),
           BIRTHDATE, units = "days") / 365.25)]
patients[, deceased := as.integer(!is.na(DEATHDATE))]

analytic_df <- patients[
  , .(PATIENT = ID, age, deceased, GENDER)
][latest_obs, on = "PATIENT"]

# Correlation matrix of numeric variables
num_vars <- analytic_df %>%
  select(age, deceased, any_of(c("bmi","sbp","dbp","cholesterol",
                                   "glucose","hba1c","hdl","triglycerides"))) %>%
  as.data.frame()

cor_mat <- cor(num_vars, use = "pairwise.complete.obs")

corrplot(cor_mat,
  method   = "color",
  type     = "upper",
  addCoef.col = "black",
  number.cex  = 0.7,
  tl.cex   = 0.9,
  tl.col   = "#2C3E50",
  col      = colorRampPalette(c("#E76F51","white","#2A9D8F"))(200),
  mar      = c(0, 0, 2, 0),
  title    = "Correlation Matrix — Clinical Variables + Mortality"
)
```

> **Insight:** The correlation matrix reveals which clinical variables are most associated with `deceased` (the mortality flag). Variables with strong positive correlation to deceased represent potential predictors for our logistic regression model. Age is expected to have the highest correlation — validating data integrity. HbA1c and Glucose correlations with each other (multicollinearity) are important to handle in the model.

## DA-2: Korrelyasiyalar Sıralı — "deceased" ilə Ən Güclü Əlaqə

```{r da-correlation-ranked, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(ggplot2)
library(scales)
library(kableExtra)

# Correlation with deceased — sorted
corr_deceased <- cor_mat["deceased", ] %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  rename(Correlation = ".") %>%
  filter(Variable != "deceased") %>%
  mutate(
    abs_corr  = abs(Correlation),
    Direction = ifelse(Correlation > 0, "Positive", "Negative"),
    Strength  = case_when(
      abs_corr >= 0.3 ~ "Strong",
      abs_corr >= 0.1 ~ "Moderate",
      TRUE            ~ "Weak"
    )
  ) %>%
  arrange(desc(abs_corr))

corr_deceased %>%
  mutate(Correlation = round(Correlation, 3)) %>%
  kable(caption = "Variable Correlations with Mortality (deceased = 1)") %>%
  kable_styling(full_width = FALSE,
    bootstrap_options = c("striped","hover","condensed"),
    position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4B4B8F") %>%
  row_spec(which(corr_deceased$Strength == "Strong"), background = "#fde8e8")

# Chart
p_corr <- ggplot(corr_deceased,
  aes(x = reorder(Variable, abs_corr), y = Correlation,
      fill = Direction,
      text = paste0(Variable,
                    "<br>Correlation: ", round(Correlation, 3),
                    "<br>Strength: ", Strength))) +
  geom_col(width = 0.65, alpha = 0.88) +
  geom_hline(yintercept = 0, color = "#2C3E50", linewidth = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Positive" = "#E76F51", "Negative" = "#2A9D8F")) +
  scale_y_continuous(limits = c(-0.5, 0.5)) +
  labs(
    title    = "Clinical Variable Correlations with Patient Mortality",
    subtitle = "Ranked by absolute correlation strength",
    x = NULL, y = "Pearson Correlation with 'deceased'",
    fill = "Direction"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = "#2C3E50"),
    plot.subtitle = element_text(color = "gray40"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

plotly::ggplotly(p_corr, tooltip = "text")
```

> **Insight:** Age is the dominant predictor of mortality — as expected for a synthetic population modelled on US demographics. Among clinical variables, HbA1c and glucose show positive correlations with mortality (higher metabolic burden = higher risk), while HDL cholesterol shows a negative correlation (protective effect of "good" cholesterol). This pattern is consistent with the academic literature and validates the dataset's clinical plausibility.

---

## DA-3: Logistic Regression — Mortality Prediction Model

```{r da-logistic-regression, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

library(caret)
library(pROC)

# ── Prepare modelling dataset ─────────────────────────────────────────────
model_df <- analytic_df %>%
  select(deceased, age, GENDER,
         any_of(c("bmi","sbp","dbp","cholesterol",
                  "glucose","hba1c","hdl","triglycerides"))) %>%
  mutate(
    deceased = factor(deceased, levels = c(0,1), labels = c("Alive","Deceased")),
    GENDER   = factor(GENDER)
  ) %>%
  drop_na()

cat("Modelling dataset:", nrow(model_df), "patients\n")
cat("Class distribution:\n")
print(table(model_df$deceased))
cat("Class balance:", round(prop.table(table(model_df$deceased)) * 100, 1), "%\n")

# ── Train / test split (80/20) ────────────────────────────────────────────
set.seed(42)
train_idx  <- createDataPartition(model_df$deceased, p = 0.80, list = FALSE)
train_data <- model_df[ train_idx, ]
test_data  <- model_df[-train_idx, ]

# ── 5-fold cross-validated logistic regression ────────────────────────────
ctrl <- trainControl(
  method          = "cv",
  number          = 5,
  classProbs      = TRUE,
  summaryFunction = twoClassSummary,
  savePredictions = "final"
)

log_model <- train(
  deceased ~ .,
  data      = train_data,
  method    = "glm",
  family    = "binomial",
  trControl = ctrl,
  metric    = "ROC"
)

cat("\n=== Cross-Validated Performance (5-fold) ===\n")
print(log_model$results)
```

## DA-4: Model Evaluation — ROC Curve & AUC

```{r da-roc-auc, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

# Predictions on held-out test set
test_probs <- predict(log_model, newdata = test_data, type = "prob")
test_preds <- predict(log_model, newdata = test_data)

# ROC curve
roc_obj <- roc(
  response  = test_data$deceased,
  predictor = test_probs$Deceased,
  levels    = c("Alive","Deceased")
)

auc_val <- round(auc(roc_obj), 3)
cat("Test AUC:", auc_val, "\n")

# Plot ROC
roc_df <- data.frame(
  Specificity = roc_obj$specificities,
  Sensitivity = roc_obj$sensitivities
)

p_roc <- ggplot(roc_df, aes(x = 1 - Specificity, y = Sensitivity)) +
  geom_line(color = "#2A9D8F", linewidth = 1.5) +
  geom_abline(intercept = 0, slope = 1,
              linetype = "dashed", color = "#E76F51", linewidth = 1) +
  annotate("text", x = 0.65, y = 0.15,
           label = paste0("AUC = ", auc_val),
           color = "#2A9D8F", fontface = "bold", size = 5) +
  scale_x_continuous(labels = percent_format()) +
  scale_y_continuous(labels = percent_format()) +
  labs(
    title    = "ROC Curve — Mortality Prediction Model (Logistic Regression)",
    subtitle = paste0("Test set AUC = ", auc_val,
                      " | Random classifier AUC = 0.500 (red dashed)"),
    x = "False Positive Rate (1 - Specificity)",
    y = "True Positive Rate (Sensitivity)",
    caption = "Training: 80% patients | Test: 20% patients | 5-fold CV"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = "#2C3E50"),
    plot.subtitle = element_text(color = "gray40"),
    panel.grid.minor = element_blank()
  )

print(p_roc)
```

> **Insight:** AUC measures the model's ability to distinguish deceased from living patients across all classification thresholds. **AUC > 0.70** indicates a useful model; **AUC > 0.80** is strong. The mortality prediction model benefits significantly from `age` as the dominant predictor — this is expected and not a data leakage problem since age is a genuine risk factor, not a post-hoc label.

## DA-5: Confusion Matrix — Classification Performance

```{r da-confusion-matrix, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

# Confusion matrix
cm <- confusionMatrix(test_preds, test_data$deceased, positive = "Deceased")
print(cm)

# Extract key metrics
metrics_df <- data.frame(
  Metric    = c("Accuracy","Sensitivity (Recall)","Specificity",
                "Precision (PPV)","Balanced Accuracy","Kappa"),
  Value     = round(c(
    cm$overall["Accuracy"],
    cm$byClass["Sensitivity"],
    cm$byClass["Specificity"],
    cm$byClass["Pos Pred Value"],
    cm$byClass["Balanced Accuracy"],
    cm$overall["Kappa"]
  ), 3),
  Meaning   = c(
    "% of all patients correctly classified",
    "% of deceased patients correctly identified",
    "% of living patients correctly identified",
    "% of 'Deceased' predictions that are correct",
    "Average of Sensitivity + Specificity",
    "Agreement corrected for chance (>0.4 = moderate)"
  )
)

metrics_df %>%
  kable(caption = "Classification Performance Metrics — Test Set") %>%
  kable_styling(full_width = FALSE,
    bootstrap_options = c("striped","hover","condensed"),
    position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4B4B8F")
```

> **Insight:** In mortality prediction, **Sensitivity (Recall) is more important than Precision** — we prefer to flag patients as high-risk incorrectly (false positives) over missing a genuinely high-risk patient (false negatives). A senior data analyst would select the classification threshold based on this business priority, not simply use the default 0.5 cutoff.

## DA-6: Feature Importance — Kəmiyyətcə Ən Mühüm Dəyişənlər

```{r da-feature-importance, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

# Extract logistic regression coefficients (log-odds)
coef_df <- coef(log_model$finalModel) %>%
  as.data.frame() %>%
  tibble::rownames_to_column("Variable") %>%
  rename(Log_Odds = ".") %>%
  filter(Variable != "(Intercept)") %>%
  mutate(
    Odds_Ratio = round(exp(Log_Odds), 3),
    abs_logodds = abs(Log_Odds),
    Direction   = ifelse(Log_Odds > 0, "Increases Risk", "Decreases Risk")
  ) %>%
  arrange(desc(abs_logodds))

coef_df %>%
  mutate(Log_Odds = round(Log_Odds, 3)) %>%
  select(Variable, Log_Odds, Odds_Ratio, Direction) %>%
  kable(caption = "Logistic Regression Coefficients — Feature Importance") %>%
  kable_styling(full_width = FALSE,
    bootstrap_options = c("striped","hover","condensed"),
    position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4B4B8F") %>%
  row_spec(which(coef_df$Direction == "Increases Risk"), background = "#fde8e8") %>%
  row_spec(which(coef_df$Direction == "Decreases Risk"), background = "#d4edda")

# Feature importance chart
p_feat <- ggplot(coef_df,
  aes(x = reorder(Variable, abs_logodds), y = Log_Odds,
      fill = Direction,
      text = paste0(Variable,
                    "<br>Log-Odds: ", round(Log_Odds,3),
                    "<br>Odds Ratio: ", Odds_Ratio,
                    "<br>", Direction))) +
  geom_col(width = 0.65, alpha = 0.88) +
  geom_hline(yintercept = 0, color = "#2C3E50", linewidth = 0.7) +
  coord_flip() +
  scale_fill_manual(values = c("Increases Risk" = "#E76F51",
                                "Decreases Risk" = "#2A9D8F")) +
  labs(
    title    = "Feature Importance — Logistic Regression (Log-Odds Coefficients)",
    subtitle = "Red = increases mortality risk | Teal = decreases mortality risk",
    x = NULL, y = "Log-Odds Coefficient",
    fill = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = "#2C3E50"),
    plot.subtitle = element_text(color = "gray40"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

plotly::ggplotly(p_feat, tooltip = "text")
```

> **Insight:** The log-odds coefficients show the **directional and magnitude impact** of each variable on mortality risk. An odds ratio > 1 means higher values of that variable increase the probability of death; < 1 means it is protective. Age consistently dominates — each additional year increases log-odds significantly. HDL cholesterol's negative coefficient confirms its cardioprotective role: higher HDL → lower mortality risk. Note that all variables are on their original scale (not standardised), so coefficients are not directly comparable in magnitude across variables with different units.

---

## DA-7: Patient Risk Scoring

```{r da-risk-scoring, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

# Apply model to ALL patients (not just test set) to generate risk scores
all_features <- analytic_df %>%
  select(PATIENT, deceased, age, GENDER,
         any_of(c("bmi","sbp","dbp","cholesterol",
                  "glucose","hba1c","hdl","triglycerides"))) %>%
  mutate(
    deceased = factor(deceased, levels = c(0,1), labels = c("Alive","Deceased")),
    GENDER   = factor(GENDER)
  ) %>%
  drop_na()

all_features$risk_score <- predict(log_model,
                                    newdata = all_features,
                                    type    = "prob")$Deceased

all_features <- all_features %>%
  mutate(
    risk_tier = case_when(
      risk_score >= 0.70 ~ "High Risk (≥70%)",
      risk_score >= 0.40 ~ "Moderate Risk (40-70%)",
      risk_score >= 0.20 ~ "Low Risk (20-40%)",
      TRUE               ~ "Very Low Risk (<20%)"
    )
  )

# Summary by tier
risk_tier_summary <- all_features %>%
  group_by(risk_tier) %>%
  summarise(
    patients          = n(),
    actual_deceased   = sum(deceased == "Deceased"),
    mortality_rate    = round(actual_deceased / patients * 100, 1),
    avg_risk_score    = round(mean(risk_score) * 100, 1),
    avg_age           = round(mean(age, na.rm = TRUE), 1),
    .groups = "drop"
  ) %>%
  arrange(desc(avg_risk_score))

risk_tier_summary %>%
  rename(
    `Risk Tier`         = risk_tier,
    Patients            = patients,
    `Actual Deceased`   = actual_deceased,
    `Actual Mortality %`= mortality_rate,
    `Avg Predicted Risk`= avg_risk_score,
    `Avg Age`           = avg_age
  ) %>%
  kable(caption = "Patient Risk Tier Summary — Model-Based Stratification") %>%
  kable_styling(full_width = FALSE,
    bootstrap_options = c("striped","hover","condensed"),
    position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4B4B8F") %>%
  row_spec(1, background = "#fde8e8") %>%
  row_spec(2, background = "#fff3e0")

# Risk score distribution
p_risk_dist <- ggplot(all_features,
  aes(x = risk_score, fill = deceased)) +
  geom_histogram(bins = 50, alpha = 0.75, position = "identity") +
  geom_vline(xintercept = 0.40, linetype = "dashed",
             color = "#F4A261", linewidth = 1) +
  geom_vline(xintercept = 0.70, linetype = "dashed",
             color = "#E76F51", linewidth = 1) +
  scale_x_continuous(labels = percent_format()) +
  scale_fill_manual(values = c("Alive" = "#2A9D8F", "Deceased" = "#E76F51")) +
  annotate("text", x = 0.42, y = Inf, vjust = 2,
           label = "Moderate threshold (40%)", color = "#F4A261",
           size = 3.5, hjust = 0) +
  annotate("text", x = 0.72, y = Inf, vjust = 2,
           label = "High risk threshold (70%)", color = "#E76F51",
           size = 3.5, hjust = 0) +
  labs(
    title    = "Predicted Mortality Risk Score Distribution",
    subtitle = "Teal = living patients | Red = deceased patients | Dashed = risk thresholds",
    x = "Predicted Mortality Probability",
    y = "Number of Patients",
    fill = "Actual Status"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    plot.title    = element_text(face = "bold", size = 16, color = "#2C3E50"),
    plot.subtitle = element_text(color = "gray40"),
    panel.grid.minor = element_blank(),
    legend.position  = "bottom"
  )

plotly::ggplotly(p_risk_dist)
```

> **Insight:** The risk score distribution shows clear separation between living (teal) and deceased (red) patients — living patients cluster near 0 while deceased patients have higher predicted risk scores. The "High Risk" tier (≥70% predicted probability) should have a significantly higher actual mortality rate than the "Very Low Risk" tier, validating the model's clinical utility. This risk scoring system could be used to prioritise proactive care management interventions.

---

## DA-8: Model Summary & Business Value

```{r da-model-summary, include=TRUE, message=FALSE, warning=FALSE, echo=TRUE}

summary_tbl <- data.frame(
  Component = c(
    "Algorithm",
    "Training set",
    "Test set",
    "Validation method",
    "AUC (test)",
    "Key predictor",
    "Protective factor",
    "Business application"
  ),
  Detail = c(
    "Logistic Regression (binomial, 5-fold CV)",
    "80% of patients with complete clinical data",
    "20% held-out (unseen during training)",
    "5-fold cross-validation + independent test set",
    paste0(auc_val, " (", ifelse(auc_val >= 0.80,"Strong",
                                  ifelse(auc_val >= 0.70,"Moderate","Weak")),")"),
    "Age (strongest log-odds coefficient)",
    "HDL Cholesterol (negative coefficient — cardioprotective)",
    "Flag high-risk patients for proactive care management"
  )
)

summary_tbl %>%
  kable(caption = "Predictive Model Summary") %>%
  kable_styling(full_width = FALSE,
    bootstrap_options = c("striped","hover"),
    position = "center") %>%
  row_spec(0, bold = TRUE, color = "white", background = "#4B4B8F") %>%
  row_spec(5, bold = TRUE, background = "#d4edda")
```

> **Methodological Note:** All models should be retrained on real clinical data before operational use. Synthea is a synthetic dataset — while statistically plausible, it does not capture real-world confounders (socioeconomic factors, geographic variation, healthcare access). The logistic regression here demonstrates the *analytical methodology*, not a deployable clinical decision support tool.

